---
title: "11_rose_plots"
format: pdf
---

### OKAY NEED TO RETHINK THE ORDER FOR ALL OF THESE PROCESSES - cannot export geometries to .csv files - the geometries are LOST ###
### Must create the geometries right before any time I need to use them ####
NEED TO COME BACK AND REVISIT THIS CODE AND OTHERS TO UPDATE IN LIGHT OF THIS



# load packages
```{r}

library(tidyverse) # includes and loads the packages listed below
# library(readr)
# library(tidyr)
# library(dplyr)
# library(ggplot2)
# library(lubridate)

library(stringr)  # for manipulating text strings

library(here) # helps with paths for files
library(arsenal)

library(purrr)

library(sf)
library(geosphere)  # for calculating change in bearing and distances for gCOBs

```

# source R scripts for functions
```{r source custom functions to simplify conducting these analyses}
#| echo: false

# getwd()

source("R_scripts/functions/fish_calculate_gCOB.R")
source("R_scripts/functions/filter_gCOB_dataframes.R")
source("R_scripts/functions/calculate_gCOB_movement.R")
source("R_scripts/functions/calculate_gCOB_annual_movement.R")

# source("R_scripts/functions/test_gCOBs_over_time.R")
# source("R_scripts/functions/combine_and_modify_models.R")
# source("R_scripts/functions/plot_lat_long_gCOB_data.R")

```



# load data subsets from .csv files
```{r load data from csv files}
#| echo: false
#| message: false

# # import csv of clean bio data
# bio_data_clean <- read_csv(file = "data/clean_data/bio_data_clean.csv")



# import csv with climate zone information for each fish species as web scraped from FishBase
species_climate_df <- read_csv(file = "output/species_climate_data.csv")


# import csv files for data subsets to explore
bio_data_clean_yr2010_2022_m101112 <- read_csv(file = "data/clean_data/bio_data_clean_yr2010_2022_m101112.csv")
bio_data_clean_yr2010_2022_m678 <- read_csv(file = "data/clean_data/bio_data_clean_yr2010_2022_m678.csv")
bio_data_clean_yr1983_2022_m678_western <-read_csv(file = "data/clean_data/bio_data_clean_yr1983_2022_m678_western.csv")

```


# IF USING DATA FROM .CSV FILES - MUST REDO GEOMETRIES

## Convert latitude and longitude into spatial data
 - Use coordinate reference system (crs) of *NAD83*, which is good for North America. It's specifically tied to the North American tectonic plate to minimize changes to coordinate values over time.
      - versus WGS84, which is global.
 - *NAD83 EPSG code = 4269*
 - WGS84 EPSG code = 4326
 - The underlying ellipsoid model is different for NAD83 versus WGS84.
      - NAD83 uses the GRS80 ellipsoid
      - WGS84 uses the WGS84 ellipsoid
- Ellipsoid parameter values
      - *NAD83*: 
          - semi-major axis ('a'): 6378137.0 meters
          - flattening ('f'): 1 / 298.257222101
      - WGS84: 
          - semi-major axis ('a'): 6378137.0 meters
          - flattening ('f'): 1 / 298.257223563

```{r}

# bio_data_clean <- st_as_sf(bio_data_clean, coords=c("DECSLON", "DECSLAT"), crs=4269, remove = FALSE)


bio_data_clean_yr2010_2022_m101112 <- st_as_sf(bio_data_clean_yr2010_2022_m101112, coords=c("DECSLON", "DECSLAT"), crs=4269, remove = FALSE)
bio_data_clean_yr2010_2022_m678 <- st_as_sf(bio_data_clean_yr2010_2022_m678, coords=c("DECSLON", "DECSLAT"), crs=4269, remove = FALSE)
bio_data_clean_yr1983_2022_m678_western <- st_as_sf(bio_data_clean_yr1983_2022_m678_western, coords=c("DECSLON", "DECSLAT"), crs=4269, remove = FALSE)

```






### entire GoM - Fall Sampling - 2010 - 2022


```{r}

```


### entire GoM - Summer Sampling - 2010 - 2022

### western GoM - Summer Sampling - 1983 - 2022

```{r}


## species list for 1983 - 2022 summer months (6,7,8) data only
species_df_yr1983_2022_m678_western <- bio_data_clean_yr1983_2022_m678_western %>% 
  select(new_TAXON, RANK) %>% 
  filter(RANK == "SPECIES") %>%
  distinct(new_TAXON, .keep_all = TRUE) %>% 
  arrange(new_TAXON)

# dim(species_vec_yr1983_2022_m678_western) # 513 unique species (649 unique taxonomic names)
species_vec_yr1983_2022_m678_western <- species_df_yr1983_2022_m678_western$new_TAXON
num_species <- length(species_vec_yr1983_2022_m678_western) # 513 unique species (649 unique taxonomic names)



# Step 1 - use fish_calculate_gCOB function
# Applying the gCOB calculate function to each unique species for the dataset only containing 2010 - 2022 data
# create an empty list to store cg dataframe
gCOBs_yr1983_2022_m678_western <- list()

# Apply the fish_calculate_gCOB function to each unique species and combine results into a list
result_list_yr1983_2022_m678_western <- purrr::map(species_vec_yr1983_2022_m678_western,
                                                   ~fish_calculate_gCOB(data = bio_data_clean_yr1983_2022_m678_western,
                                                                        output_name = "gCOBs_yr1983_2022_m678_western",
                                                                        taxonomic_column = "new_TAXON",
                                                                        common_name_column = "new_TAXON",
                                                                        species_name = .))



# Step 2 - filter to only include fish species with 5+ gCOB observations
# Filter the list of data frames
gCOBs_yr1983_2022_m678_western <- filter_gCOB_dataframes(gCOBs_yr1983_2022_m678_western, min_years = 5)


# Step 3 - calculate change in gCOBs over time in terms of bearing and distance

#initiate empty list
movement_results <- list()

gCOBs_data <- get("gCOBs_yr1983_2022_m678_western", envir = .GlobalEnv)
movement_results <- calculate_all_gCOB_movements(gCOBs_data)


# Step 3 part 2 - calculate change in gCOBs from year to year (bearing and distance) for each species

gCOBs_data <- get("gCOBs_all_data", envir = .GlobalEnv)
annual_movement_results <- calculate_annual_gCOB_movements(gCOBs_data)



```

###### make wind rose plots

```{r}

# Bin the bearing values and aggregate data
binned_climate_gCOBs <- movement_results %>%
  rename(species_name = species) %>%
  left_join(species_climate_df, join_by(species_name)) %>% 
  rename(species = species_name) %>% 
  mutate(bearing_bins = cut(bearing, breaks = seq(0, 360, length.out = 37), include.lowest = TRUE)) %>% 
  group_by(climate_zone, bearing_bins) %>% 
  summarize(num_species = n(), .groups = 'drop') %>% 
  ungroup()

# Extract labels for bins based on their midpoints
bin_labels <- seq(0, 360, by = 10) # Midpoints of 36 bins (0-10, 10-20, ..., 350-360)

# sequence of labels that are not every 10 degrees
less_bin_labels <- c("0", rep("",8), "90", rep("",8), "180", rep("",8), "270", rep("",8))

# Ensure the correct order of bearing bins
binned_climate_gCOBs <- binned_climate_gCOBs %>%
  mutate(bearing_bins = factor(bearing_bins,
                               levels = levels(bearing_bins),
                               labels = bin_labels[1:36]))


# make plot
ggplot(data = binned_climate_gCOBs, aes(x = bearing_bins, y = num_species,
                                        fill = climate_zone,
                                        group = climate_zone)) +
  geom_bar(stat = "identity", width = 1, color = "black") +
  coord_polar(theta = "x", start = -pi/32, direction = 1) +   # adjust start for orientation
  scale_fill_manual(values = c("tropical" = "coral", "temperate" = "skyblue", "subtropical" = "lightgreen", "deep-water" = "navy")) +
  scale_y_continuous(expand = c(0, 0)) +
  scale_x_discrete(drop = FALSE,
    # limits = as.character(bin_labels[1:36]),
    # limits = levels(bearing_bins),
                   # labels = bin_labels[1:36],    # Apply bin midpoints as labels
                   labels = less_bin_labels,    # less labels
                   expand = c(0,0.5)
                   ) +  
  geom_text(aes(label = num_species), position = position_stack(vjust = 0.9), size = 3) +  # Add labels to petals
  # theme_minimal() +
  # theme(axis.title = element_blank(),
  #       axis.text.x = element_text(size = 12, vjust = 0.5),
  #       axis.text.y = element_blank(),
  #       panel.grid = element_blank(),
  #       legend.position = "right") +
  labs(fill = "Climate Distribution") +
  ggtitle("Wind Rose Plot with Species Counts and Climate Distribution") +
  facet_wrap(~climate_zone, scales = "free_y")

```


## wind rose plots showing averages

```{r}


# # Summarize gCOB data by species: average bearing, distance, and retain sample size, earliest, and latest year
# summarized_gCOBs <- annual_movement_results %>%
#   group_by(species) %>%
#   summarize(
#     avg_bearing = mean(bearing, na.rm = TRUE),
#     avg_distance = mean(distance, na.rm = TRUE),
#     sample_size = n(),
#     earliest_year = min(as.numeric(substring(year_pair, 1, 4)), na.rm = TRUE),
#     latest_year = max(as.numeric(substring(year_pair, 6, 9)), na.rm = TRUE)
#   ) %>%
#   ungroup()


# Function to calculate standard error
std_error <- function(x) {
  sd(x, na.rm = TRUE) / sqrt(sum(!is.na(x)))
}

# Summarize gCOB data by species: calculate mean, SD, and SE for bearing and distance
summarized_gCOBs <- annual_movement_results %>%
  group_by(species) %>%
  summarize(
    avg_bearing = mean(bearing, na.rm = TRUE),
    sd_bearing = sd(bearing, na.rm = TRUE),  # Standard deviation of bearing
    se_bearing = std_error(bearing),  # Standard error of bearing
    avg_distance = mean(distance, na.rm = TRUE),
    sd_distance = sd(distance, na.rm = TRUE),  # Standard deviation of distance
    se_distance = std_error(distance),  # Standard error of distance
    sample_size = n(),
    earliest_year = min(as.numeric(substring(year_pair, 1, 4)), na.rm = TRUE),
    latest_year = max(as.numeric(substring(year_pair, 6, 9)), na.rm = TRUE)
  ) %>%
  ungroup()



# Join with climate zone data
binned_climate_gCOBs <- summarized_gCOBs %>%
  rename(species_name = species) %>%
  left_join(species_climate_df, join_by(species_name)) %>% 
  rename(species = species_name) %>%
  # Bin the average bearing into 36 bins
  mutate(bearing_bins = cut(avg_bearing, breaks = seq(0, 360, length.out = 37), include.lowest = TRUE)) %>%
  group_by(climate_zone, bearing_bins) %>%
  summarize(
    num_species = n(),
    avg_distance = mean(avg_distance, na.rm = TRUE),
    .groups = 'drop'
  ) %>%
  ungroup()

# Extract labels for bins based on their midpoints
bin_labels <- seq(0, 360, by = 10) # Midpoints of 36 bins (0-10, 10-20, ..., 350-360)

# Labels for key angles (0, 90, 180, 270)
less_bin_labels <- c("0", rep("",8), "90", rep("",8), "180", rep("",8), "270", rep("",8))

# Ensure the correct order of bearing bins
binned_climate_gCOBs <- binned_climate_gCOBs %>%
  mutate(bearing_bins = factor(bearing_bins,
                               levels = levels(bearing_bins),
                               labels = bin_labels[1:36]))

# Create the wind rose plot
ggplot(data = binned_climate_gCOBs, aes(x = bearing_bins, y = num_species,
                                        fill = climate_zone,
                                        group = climate_zone)) +
  geom_bar(stat = "identity", width = 1, color = "black") +
  coord_polar(theta = "x", start = -pi/32, direction = 1) +   # Adjust start for correct orientation (due north at 12 o'clock)
  scale_fill_manual(values = c("tropical" = "coral", "temperate" = "skyblue", "subtropical" = "lightgreen", "deep-water" = "navy")) +
  scale_y_continuous(expand = c(0, 0)) +
  scale_x_discrete(drop = FALSE,
                   labels = less_bin_labels,    # Less labels for 0, 90, 180, 270 degrees
                   expand = c(0,0.5)) +  
  geom_text(aes(label = num_species), position = position_stack(vjust = 0.9), size = 3) +  # Add labels to petals
  theme_minimal() +  # Apply a minimal theme for a clean look
  theme(axis.title = element_blank(),
        axis.text.x = element_text(size = 12, vjust = 0.5),
        axis.text.y = element_blank(),
        panel.grid = element_blank(),
        legend.position = "right") +  # Position the legend to the right
  labs(fill = "Climate Distribution") +
  ggtitle("Wind Rose Plot with Averaged gCOBs and Climate Distribution") +
  facet_wrap(~climate_zone, scales = "free_y")  # Facet by climate zone with free y-scales

```


# wind rose plot for just tropical fish species - showing averages and error bars
```{r}

# Filter summarized data for only tropical species
tropical_gCOBs <- summarized_gCOBs %>%
  rename(species_name = species) %>%
  left_join(species_climate_df, join_by(species_name)) %>% 
  rename(species = species_name) %>%
  filter(climate_zone == "tropical") %>%
  drop_na(avg_bearing, se_bearing)  # Ensure no missing values in avg_bearing or se_bearing



# Create the bar plot for tropical species
ggplot(tropical_gCOBs, aes(x = species, y = avg_bearing)) +
  geom_bar(stat = "identity", width = 1, fill = "coral", color = "black") +
  geom_errorbar(aes(ymin = avg_bearing - se_bearing, ymax = avg_bearing + se_bearing), width = 0.2) +  # Error bars
  # coord_polar(theta = "y", start = 0) +  # Polar plot to represent wind rose
  scale_y_continuous(limits = c(0, 360), expand = c(0, 0)) +  # Limit bearing from 0 to 360 degrees
  theme_minimal() +  # Minimal theme for a clean look
  theme(
    axis.title = element_blank(),  # No axis titles
    axis.text.x = element_text(size = 10, vjust = 0.5, angle = 90),
    # axis.text.y = element_blank(),  # Remove y-axis labels
    panel.grid = element_blank()  # Remove grid lines for clarity
  ) +
  labs(title = "Wind Rose of Tropical Fish Species", y = "Average Bearing (degrees)") +
  ggtitle("Wind Rose Plot: Tropical Fish Species Bearing with Error Bars")



```



```{r}

# Create the wind rose plot
ggplot(tropical_gCOBs, aes(x = avg_bearing, y = avg_distance, fill = species)) +
  geom_bar(stat = "identity", width = 10, color = "black") +  # Bar width represents bearing intervals
  geom_errorbar(aes(x = avg_bearing, ymin = avg_distance - se_distance, ymax = avg_distance + se_distance), 
                width = 0.2, color = "red") +  # Error bars for distance
  coord_polar(theta = "x", start = -pi/2, direction = 1) +  # Polar plot to represent wind rose
  scale_x_continuous(breaks = seq(0, 360, by = 30),  # Bearings at every 30 degrees
                     limits = c(0, 360), 
                     expand = c(0, 0)) +
  scale_y_continuous(expand = c(0, 0)) +  # Adjust y-axis to start at 0
  # scale_fill_manual(values = "coral") +  # Single color for tropical species
  theme_minimal() +  # Clean theme
  theme(
    axis.title.y = element_blank(),
    axis.text.y = element_blank(),  # Remove radial axis labels
    panel.grid = element_blank(),   # Remove grid lines
    legend.position = "none"        # No need for a legend if plotting one species type
  ) +
  labs(title = "Wind Rose Plot: Tropical Fish Species", x = "Bearing (degrees)", y = "Distance") +
  ggtitle("Wind Rose Plot: Average Bearing and Distance for Tropical Fish Species")


```









## CODE FROM CHATGPT - NEEDS TO BE ADAPTED TO MAKE A ROSE PLOT

```{r}


# Load necessary libraries
library(ggplot2)
library(dplyr)
library(scales)

# Sample data (replace this with your actual data)
set.seed(123)
data <- data.frame(
  compass_bearing = sample(seq(0, 360, by = 10), 100, replace = TRUE),
  distance = runif(100, 1, 10),
  climate_distribution = sample(c("Tropical", "Temperate", "Arctic"), 100, replace = TRUE)
)

# Create the wind rose plot
ggplot(data, aes(x = factor(compass_bearing), y = distance, fill = climate_distribution)) +
  geom_bar(stat = "identity", width = 1, color = "black") +
  coord_polar(start = -pi/16) +  # Adjust start for orientation
  scale_fill_manual(values = c("Tropical" = "coral", "Temperate" = "skyblue", "Arctic" = "lightgreen")) +
  scale_y_continuous(expand = c(0, 0)) +
  theme_minimal() +
  theme(axis.title = element_blank(),
        axis.text.x = element_text(size = 12, vjust = 0.5),
        axis.text.y = element_blank(),
        panel.grid = element_blank(),
        legend.position = "right") +
  labs(fill = "Climate Distribution") +
  ggtitle("Wind Rose Plot with Distance and Climate Distribution")


```



