---
title: "11_rose_plots"
format: pdf
---

### OKAY NEED TO RETHINK THE ORDER FOR ALL OF THESE PROCESSES - cannot export geometries to .csv files - the geometries are LOST ###
### Must create the geometries right before any time I need to use them ####
NEED TO COME BACK AND REVISIT THIS CODE AND OTHERS TO UPDATE IN LIGHT OF THIS



# load packages
```{r}

library(tidyverse) # includes and loads the packages listed below
# library(readr)
# library(tidyr)
# library(dplyr)
# library(ggplot2)
# library(lubridate)

library(stringr)  # for manipulating text strings

library(here) # helps with paths for files
library(arsenal)

library(purrr)

library(sf)
library(geosphere)  # for calculating change in bearing and distances for gCOBs

```


```{r source custom functions to simplify conducting these analyses}
#| echo: false

# getwd()

source("R_scripts/functions/fish_calculate_gCOB.R")
source("R_scripts/functions/filter_gCOB_dataframes.R")
source("R_scripts/functions/calculate_gCOB_movement.R")

# source("R_scripts/functions/test_gCOBs_over_time.R")
# source("R_scripts/functions/combine_and_modify_models.R")
# source("R_scripts/functions/plot_lat_long_gCOB_data.R")

```




```{r load data from csv files}
#| echo: false
#| message: false

# # import csv of clean bio data
# bio_data_clean <- read_csv(file = "data/clean_data/bio_data_clean.csv")
# 
# # import csv with climate zone information for each fish species as web scraped from FishBase
# species_climate_df <- read_csv(file = "output/species_climate_data.csv")


# import csv files for data subsets to explore
bio_data_clean_yr2010_2022_m101112 <- read_csv(bio_data_clean_yr2010_2022_m101112, file = "data/clean_data/bio_data_clean_yr2010_2022_m101112.csv")
bio_data_clean_yr2010_2022_m678 <- read_csv(bio_data_clean_yr2010_2022_m678, file = "data/clean_data/bio_data_clean_yr2010_2022_m678.csv")
bio_data_clean_yr1983_2022_m678_western <-read_csv(bio_data_clean_yr1983_2022_m678_western, file = "data/clean_data/bio_data_clean_yr1983_2022_m678_western.csv")

```


### entire GoM - Fall Sampling - 2010 - 2022

### entire GoM - Summer Sampling - 2010 - 2022

### western GoM - Summer Sampling - 1983 - 2022

```{r}


## species list for 2010 - 2022 fall months (10,11,12) data only
species_df_yr1983_2022_m678_western <- bio_data_clean_yr1983_2022_m678_western %>% 
  select(new_TAXON, RANK) %>% 
  # filter(RANK == "SPECIES") %>% 
  distinct(new_TAXON, .keep_all = TRUE) %>% 
  arrange(new_TAXON)

# dim(species_df_yr2010_2022_m101112) # 500 unique taxonomic names
species_vec_yr1983_2022_m678_western <- species_df_yr1983_2022_m678_western$new_TAXON
num_species <- length(species_vec_yr1983_2022_m678_western) # 500 unique taxonomic names



# Step 1 - use fish_calculate_gCOB function
# Applying the gCOB calculate function to each unique species for the dataset only containing 2010 - 2022 data
# create an empty list to store cg dataframe
gCOBs_yr1983_2022_m678_western <- list()

# Apply the fish_calculate_gCOB function to each unique species and combine results into a list
result_list_yr1983_2022_m678_western <- purrr::map(species_vec_yr1983_2022_m678_western,
                                                   ~fish_calculate_gCOB(data = bio_data_clean_yr1983_2022_m678_western,
                                                                        output_name = "gCOBs_yr1983_2022_m678_western",
                                                                        taxonomic_column = "new_TAXON",
                                                                        common_name_column = "new_TAXON",
                                                                        species_name = .))



# Step 2 - filter to only include fish species with 5+ gCOB observations
# Filter the list of data frames
gCOBs_yr1983_2022_m678_western <- filter_gCOB_dataframes(gCOBs_yr1983_2022_m678_western, min_years = 5)


# Step 3 - calculate change in gCOBs over time in terms of bearing and distance

#initiate empty list
movement_results <- list()

gCOBs_data <- get("gCOBs_yr1983_2022_m678_western", envir = .GlobalEnv)
movement_results <- calculate_all_gCOB_movements(gCOBs_data)


```

###### make wind rose plots

```{r}

# Bin the bearing values and aggregate data
binned_climate_gCOBs <- climate_gCOB_movements %>% 
  mutate(bearing_bins = cut(bearing, breaks = seq(0, 360, length.out = 37), include.lowest = TRUE)) %>% 
  group_by(climate_zone, bearing_bins) %>% 
  summarize(num_species = n(), .groups = 'drop') %>% 
  ungroup()

# Extract labels for bins based on their midpoints
bin_labels <- seq(0, 360, by = 10) # Midpoints of 36 bins (0-10, 10-20, ..., 350-360)

# sequence of labels that are not every 10 degrees
less_bin_labels <- c("0", rep("",8), "90", rep("",8), "180", rep("",8), "270", rep("",8))

# Ensure the correct order of bearing bins
binned_climate_gCOBs <- binned_climate_gCOBs %>%
  mutate(bearing_bins = factor(bearing_bins,
                               levels = levels(bearing_bins),
                               labels = bin_labels[1:36]))


# make plot
ggplot(data = binned_climate_gCOBs, aes(x = bearing_bins, y = num_species,
                                        fill = climate_zone,
                                        group = climate_zone)) +
  geom_bar(stat = "identity", width = 1, color = "black") +
  coord_polar(theta = "x", start = -pi/32, direction = 1) +   # adjust start for orientation
  scale_fill_manual(values = c("tropical" = "coral", "temperate" = "skyblue", "subtropical" = "lightgreen", "deep-water" = "navy")) +
  scale_y_continuous(expand = c(0, 0)) +
  scale_x_discrete(drop = FALSE,
    # limits = as.character(bin_labels[1:36]),
    # limits = levels(bearing_bins),
                   # labels = bin_labels[1:36],    # Apply bin midpoints as labels
                   labels = less_bin_labels,    # less labels
                   expand = c(0,0.5)
                   ) +  
  geom_text(aes(label = num_species), position = position_stack(vjust = 0.9), size = 3) +  # Add labels to petals
  # theme_minimal() +
  # theme(axis.title = element_blank(),
  #       axis.text.x = element_text(size = 12, vjust = 0.5),
  #       axis.text.y = element_blank(),
  #       panel.grid = element_blank(),
  #       legend.position = "right") +
  labs(fill = "Climate Distribution") +
  ggtitle("Wind Rose Plot with Species Counts and Climate Distribution") +
  facet_wrap(~climate_zone, scales = "free_y")

```
















## CODE FROM CHATGPT - NEEDS TO BE ADAPTED TO MAKE A ROSE PLOT

```{r}


# Load necessary libraries
library(ggplot2)
library(dplyr)
library(scales)

# Sample data (replace this with your actual data)
set.seed(123)
data <- data.frame(
  compass_bearing = sample(seq(0, 360, by = 10), 100, replace = TRUE),
  distance = runif(100, 1, 10),
  climate_distribution = sample(c("Tropical", "Temperate", "Arctic"), 100, replace = TRUE)
)

# Create the wind rose plot
ggplot(data, aes(x = factor(compass_bearing), y = distance, fill = climate_distribution)) +
  geom_bar(stat = "identity", width = 1, color = "black") +
  coord_polar(start = -pi/16) +  # Adjust start for orientation
  scale_fill_manual(values = c("Tropical" = "coral", "Temperate" = "skyblue", "Arctic" = "lightgreen")) +
  scale_y_continuous(expand = c(0, 0)) +
  theme_minimal() +
  theme(axis.title = element_blank(),
        axis.text.x = element_text(size = 12, vjust = 0.5),
        axis.text.y = element_blank(),
        panel.grid = element_blank(),
        legend.position = "right") +
  labs(fill = "Climate Distribution") +
  ggtitle("Wind Rose Plot with Distance and Climate Distribution")


```



