---
title: "1_organize_data"
author: "Shayna A. Sura"
format: html
editor: source
---

# Organize SEAMAP Trawl Data for Analysis

### load necessary packages

```{r load packages}
#| echo: false

library(tidyverse) # includes and loads the packages listed below
# library(readr)
# library(tidyr)
# library(dplyr)
# library(ggplot2)
# library(lubridate)

library(here) # helps with paths for files
library(arsenal)

# library(scales)
# library(ggbeeswarm)

```

### function to load multiple .csv files at once

```{r function to load csv files}
#| echo: false

mass.csv.in<-function(path){

  #Retrieving the file names of csv files in designated path
  temp<-dir(path=path, pattern="*.csv", full.names = FALSE) %>% substr(0,nchar(.)-4)

  #Subtracting files from the list which are already loaded
  temp<-setdiff(dir(path = path, pattern = ".csv"), ls(.GlobalEnv))

  #Return message if no files need to be loaded
  if(length(temp)==0){return("ALL FILES ALREADY LOADED")}

  #Loading files
  for (i in 1:length(temp)) {
    assign(temp[i], read_csv(paste(path,temp[i], sep = "")), pos = ".GlobalEnv")
    print("Files Loaded")
  }
}

```

### load SEAMAP trawl data

```{r load SEAMAP trawl data from .csv files}
#| echo: false

mass.csv.in(path = "data/seamap_trawls/")

```

### exploring SEAMAP trawl data files

```{r data exploration}
#| echo: false
#| eval: false


glimpse(BGSREC.csv)
summary(BGSREC.csv)

unique(STAREC.csv$HAULVALUE)
summary(STAREC.csv$HAULVALUE)


which(STAREC.csv$HAULVALUE != "G")
length(which(STAREC.csv$HAULVALUE != "G"))


length(which(STAREC.csv$HAULVALUE == "B"))
length(which(STAREC.csv$HAULVALUE == "G"))



head()
unique()
colnames()

```

## Cleaning the SEAMAP Trawl Data

1.  NEWBIOCODESBIG.csv file has 2 biocodes/codes that are duplicated for different taxa.

-   CODES: 170026003 & 300000000
-   I am ignoring the 300000000 code because that's for invertebrates and not applicable for my research question.
-   I am changing one occurrence of 17002600***3*** to 17002600***4*** to match the species scientific name with the other occurrence of 170026004.
    -   170026003 is listed for both ANTHIAS WOODSI and ANTHIAS TENUIS
    -   170026004 is also listed for ANTHIAS WOODSI, but there's no other biocode associated with ANTHIAS TENUIS.
-   So I am changing the occurrence of 170026003 that says ANTHIAS WOODSI to the biocode 170026004.
-   Also, the other occurrence of ANTHIAS WOODSI does not have a common name associated with it, but there is a common name, so I'm going to add that in so both occurrences of ANTHIAS WOODSI will match now. I am also changing the Key1 ID values to the first Key1 value associated with ANTHIAS WOODSI, which is Key1 = 504.
-   I am making these changes to a new data frame (taxa_names), and keeping the NEWBIOCODESBIG.csv file AS IS.

```{r taxanomic names data cleaning}

# make a new df to easily join with the biological data and to remove extra columns I don't need.
taxa_names <- NEWBIOCODESBIG.csv %>% 
  select(Key1, TAXONOMIC, CODE, common_name) %>% 
  rename(BIO_BGS = CODE)


# check for duplicated entries for the Biocodes, which will create extra entries when I join the scientific names to the biological data
which(duplicated(taxa_names$BIO_BGS))
# 275 and 10015 are duplicated BIO_BGS codes in the taxa_names dataframe


# check the rows that have duplicates to retrieve the Biocodes for these duplicated rows
taxa_names[275,] #BIO_BGS is 170026003
taxa_names[10015,] #BIO_BGS is 300000000

# find all the entries with the biocodes that are duplicated and check surrounding biocodes 
filter(taxa_names, BIO_BGS == 170026003)
filter(taxa_names, BIO_BGS == 300000000) # this is for inverts so I'm going to ignore this now.

# find all the entries with the two scientific names for this biocode to see if there's another biocode associated with either of these species
filter(taxa_names, TAXONOMIC == "ANTHIAS WOODSI")
filter(taxa_names, TAXONOMIC == "ANTHIAS TENUIS")
# ANTHIAS WOODSI has 2 different biocodes assigned to it, while ANTHIS TENUIS only has 1.

# check the surrounding biocodes (as another way to detect what's being duplicated and how to correct it)
filter(taxa_names, BIO_BGS == 170026004)
filter(taxa_names, BIO_BGS == 170026002)

# change the entry for ANTHIAS WOODSI that has the wrong biocode associated with it
taxa_names <- taxa_names %>% 
  mutate(BIO_BGS = case_when((BIO_BGS == 170026003 & TAXONOMIC == "ANTHIAS WOODSI") ~ 170026004, .default = BIO_BGS),
         common_name = case_when((BIO_BGS == 170026004 & TAXONOMIC == "ANTHIAS WOODSI") ~ "swallowtail bass", .default = common_name),
         Key1 = case_when((BIO_BGS == 170026004 & TAXONOMIC == "ANTHIAS WOODSI") ~ 504, .default = Key1))

# double checking that the above mutate worked as expected.
filter(taxa_names, TAXONOMIC == "ANTHIAS WOODSI")
filter(taxa_names, BIO_BGS == 170026003)


# check for duplicated entries for the Biocodes, which will create extra entries when I join the scientific names to the biological data
which(duplicated(taxa_names$BIO_BGS))
# 10015 and 10331 are duplicated BIO_BGS codes in the taxa_names dataframe

taxa_names[10331,] #this is the updated entry for ANTHIAS WOODSI, which is now duplicated because we made the two entries match each other exactly. This will be okay for the join below.

```

```{r subsetting CRUISES and STAREC data}

# subset CRUISES
year_data <- CRUISES.csv %>% 
  select(CRUISEID, YR)

# checking the values in the YR column to make sure there aren't any weird ones that might need to be fixed
unique(year_data$YR)
length(unique(year_data$YR))

# subset STAREC
small_starec_data <- STAREC.csv %>% 
  select(STATIONID, CRUISEID, CRUISE_NO, DEPTH_SSTA, MO_DAY_YR, DECSLAT, DECSLON, DECELAT, DECELON)

```

## Organizing SEAMAP Trawl Data

1.  Filter out the invertebrates - keep only those data with a CATEGORY value of 3.
2.  Add in the full scientific name for the fishes collected through the BIO_BGS variable in BGSREC to the CODE variable in the NEWBIOCODESBIG.csv and pull the TAXONOMIC and the Common_name variables.
3.  Link the data to year it was collected through the CRUISEID and YR variable in CRUISES.csv
4.  Link the data to a specific date it was collected through the CRUISEID and MO_DAY_YR variable in the STAREC.csv
5.  Link the data to the HAULVALUE code from the STAREC.csv file.

```{r data organizing}


# 1. filter out invertebrates and remove extra columns I don't need from BGSREC.csc
bio_data <- BGSREC.csv %>% 
  filter(CATEGORY == 3) %>% 
  select(-CNT, -SAMPLE_BGS, -NODC_BGS, -TAXONID, -INVRECID)

# 2. add the full scientific name for the fishes
bio_data <- bio_data %>% 
  left_join(taxa_names, join_by(BIO_BGS))

# # check the bio_data df for the duplicated biocodes from above (ensure the issue was resolved)
# filter(bio_data, BIO_BGS == 300000000)
# filter(bio_data, BIO_BGS == 170026003)
# filter(bio_data, BIO_BGS == 170026004) # interesting that this fish species wasn't even found in any of the trawls.
# filter(bio_data, GENUS_BGS == "ANTHIAS")


# 3. and 4. add the year for the biological catch data and add the specific month and day.
bio_data <- bio_data %>% 
  left_join(year_data, join_by(CRUISEID)) %>% 
  left_join(small_starec_data, join_by(STATIONID))


## check that the CRUISEID and CRUISE_NO columns from the bio_data and the small_starec_data match
length(which((bio_data$CRUISEID.x == bio_data$CRUISEID.y) == "TRUE"))
length(which((bio_data$CRUISEID.x == bio_data$CRUISEID.y) == "FALSE")) # these 2 columns match

## the CRUISE_NO variable from STAREC.csv -> small_starec_data was stored as character values instead of numeric. Need to convert the CRUISE_NO variable to numeric for it to match the bio_data CRUISE_NO variable.
length(which((bio_data$CRUISE_NO.x == as.numeric(bio_data$CRUISE_NO.y)) == "TRUE"))
length(which((bio_data$CRUISE_NO.x == as.numeric(bio_data$CRUISE_NO.y)) == "FALSE")) # these 2 columns match

## remove the extra CRUISEID and CRUISE_NO columns to streamline data table
bio_data <- bio_data %>% 
  select(-CRUISEID.y, -CRUISE_NO.y) %>% 
  rename(CRUISEID = CRUISEID.x) %>% 
  rename(CRUISE_NO = CRUISE_NO.x)

## relocate columns in bio_data to have important variables towards the left side of the data table
# bio_data <- bio_data %>% 
#   relocate()


# 5. add the HAULVALUE codes
which(bio_data$STATIONID == 173641)
# the row in STAREC.csv that wasn't loaded properly into R is not in my biological data set, so I'm not going to worry about this.

haul_values <- STAREC.csv %>% 
  select(STATIONID, HAULVALUE)

bio_data <- bio_data %>% 
  left_join(haul_values, join_by(STATIONID))

unique(bio_data$HAULVALUE)
length(which(bio_data$HAULVALUE == "G"))
length(which(bio_data$HAULVALUE == "B"))
length(which(bio_data$HAULVALUE == "NA"))

bad_hauls <- bio_data %>% 
  filter(HAULVALUE != "G")

bad_hauls_2 <- bio_data %>% 
  filter(HAULVALUE != "G" | is.na(HAULVALUE) == TRUE)



# head(bio_data)

```

### BAD TRAWLS in Biological Data

1986 (29 data points)

-   CRUISEID = 76
-   STATIONID = 85594
-   CRUISE_NO = 160
-   P_STA_NO = 44108

2005 (2 data points)

-   CRUISEID = 632
-   STATIONID = 82569
-   CRUISE_NO = 267
-   P_STA_NO = 90

2005 (10 data points)

-   CRUISEID = 626
-   TATIONID = 82707
-   CRUISE_NO = 55
-   P_STA_NO = 41



```{r export clean bio_data}

write_csv(bio_data, file = "data/clean_data/bio_data.csv")

getwd()

```




# Visualize Data

```{r}

length(unique(bio_data$TAXONOMIC)) # 1059 unique taxonomic values
length(unique(bio_data$GENUS_BGS)) # 487 unique genus values


# sort(unique(bio_data$TAXONOMIC)) # taking a peek to see if there are problems with things being duplicated from misspellings - does not appear to be the case.


# try things with a subset of data...

anchoa_data <- bio_data %>% 
  filter(GENUS_BGS == "ANCHOA")


ggplot(anchoa_data, aes(x = YR, y = CNTEXP, color = TAXONOMIC)) +
  geom_line() +
  facet_wrap(~TAXONOMIC)

```




# Visualize Data - Sampling Locations

```{r}

library(leaflet)


# how can I filter out the excess data from bio_data so I only have one set of coordinates for each trawl station?
## - the small_starec_data only has 1 row for each STATIONID.

trawl_locations <- inner_join(x = small_starec_data, y = bio_data, by = join_by(STATIONID))
# hmm this is still giving me ALL the rows....
# need to select out the unnecessary columns from bio_data?


trawl_locations_2 <- (small_starec_data$STATIONID == bio_data$STATIONID) == TRUE # this doesn't work...


leaflet(bio_data)

```






