---
title: "1_organize_data"
author: "Shayna A. Sura"
format: html
editor: source
---

# Organize SEAMAP Trawl Data for Analysis

### load necessary packages

```{r load packages}
#| echo: false

library(readr)
library(here)
library(tidyr)
library(dplyr)
library(ggplot2)

# library(scales)
# library(ggbeeswarm)

```



### function to load multiple .csv files at once
```{r function to load csv files}
#| echo: false

mass.csv.in<-function(path){

  #Retrieving the file names of csv files in designated path
  temp<-dir(path=path, pattern="*.csv", full.names = FALSE) %>% substr(0,nchar(.)-4)

  #Subtracting files from the list which are already loaded
  temp<-setdiff(dir(path = path, pattern = ".csv"), ls(.GlobalEnv))

  #Return message if no files need to be loaded
  if(length(temp)==0){return("ALL FILES ALREADY LOADED")}

  #Loading files
  for (i in 1:length(temp)) {
    assign(temp[i], read_csv(paste(path,temp[i], sep = "")), pos = ".GlobalEnv")
    print("Files Loaded")
  }
}

```



### load SEAMAP trawl data

```{r load SEAMAP trawl data from .csv files}
#| echo: false

mass.csv.in(path = "data/seamap_trawls/")

```


### exploring SEAMAP trawl data files

```{r data exploration}
#| echo: false
#| eval: false


glimpse(BGSREC.csv)
summary(BGSREC.csv)

unique(STAREC.csv$HAULVALUE)
summary(STAREC.csv$HAULVALUE)


which(STAREC.csv$HAULVALUE != "G")
length(which(STAREC.csv$HAULVALUE != "G"))


length(which(STAREC.csv$HAULVALUE == "B"))
length(which(STAREC.csv$HAULVALUE == "G"))



head()
unique()
colnames()

```



## Cleaning the SEAMAP Trawl Data

1. Filter out the invertebrates
 - keep only those data with a CATEGORY value of 3
2. Link the data to year it was collected through the CRUISEID and YR variable in CRUISES.csv
3. Link the data to a specific date it was collected through the CRUISEID and MO_DAY_YR variable in the STAREC.csv
4. Add in the full scientific name for the fishes collected through the BIO_BGS variable in BGSREC to the CODE variable in the NEWBIOCODESBIG.csv and pull the TAXONOMIC and the Common_name variables.
5. 


```{r data organizing}

bio_data <- BGSREC.csv %>% 
  filter(CATEGORY == 3)

head(bio_data)

```


