---
title: "1a_data_subsets"
author: "Shayna A. Sura"
format: html
---






```{r load data from csv file and filter for desired years and months}
#| echo: false
#| message: false

# import csv of clean bio data
bio_data_clean <- read_csv(file = "data/clean_data/bio_data_clean.csv")

# import csv with climate zone information for each fish species as web scraped from FishBase
species_climate_df <- read_csv(file = "output/species_climate_data.csv")


```


# join climate data to bio_data_clean

```{r}

# join climate zone information to gCOB_trends
bio_data_clean <- bio_data_clean %>%
  rename(species_name = new_TAXON) %>%
  left_join(species_climate_df, join_by(species_name)) %>% 
  rename(new_TAXON = species_name)

```


## subset data for All of Gulf of Mexico for Fall Sampling: 2010 - 2022
```{r}


# subset the data for the target years and months
years_to_include <- c(2010:2022)
months_to_include <- c(10,11,12)


bio_data_clean_yr2010_2022_m101112 <- bio_data_clean %>% 
  filter(YR %in% years_to_include) %>% 
  mutate(MONTH = month(dmy(MO_DAY_YR))) %>% 
  filter(MONTH %in% months_to_include)


# export data subset to csv file
write_csv(bio_data_clean_yr2010_2022_m101112, file = "data/clean_data/bio_data_clean_yr2010_2022_m101112.csv")

```


## subset data for All of Gulf of Mexico for Summer Sampling: 2010 - 2022
```{r}

# subset the data for the target years and months
years_to_include <- c(2010:2022)
months_to_include <- c(6,7,8)


bio_data_clean_yr2010_2022_m678 <- bio_data_clean %>% 
  filter(YR %in% years_to_include) %>% 
  mutate(MONTH = month(dmy(MO_DAY_YR))) %>% 
  filter(MONTH %in% months_to_include)


# export data subset to csv file
write_csv(bio_data_clean_yr2010_2022_m678, file = "data/clean_data/bio_data_clean_yr2010_2022_m678.csv")

```


########### NEED TO UPDATE THIS FILTERING BELOW TO USE THE GEOMETRIES FOR FILTERING INSTEAD OF THE DESCLON ###########
## Or at least investigate if it makes much of a difference

## subset data for western Gulf of Mexico for summer trawls (June, July, August) in years 1983 - 2022
```{r}

# subset the data for the target years and months
years_to_include <- c(1983:2022)
months_to_include <- c(6,7,8)


bio_data_clean_yr1983_2022_m678_western <- bio_data_clean %>% 
  filter(YR %in% years_to_include) %>% 
  mutate(MONTH = month(dmy(MO_DAY_YR))) %>% 
  filter(MONTH %in% months_to_include) %>% 
  filter(DECSLON < -87.4)


# max(bio_data_clean_yr1983_2022_m678_western$DECSLON)


# export data subset to csv file
write_csv(bio_data_clean_yr1983_2022_m678_western, file = "data/clean_data/bio_data_clean_yr1983_2022_m678_western.csv")

```


