---
title: "2_AFS_station_maps"
author: "Shayna A. Sura"
format: html
---

## Make animated map of trawl station locations over time

### load packages
```{r load packages}
#| echo: false
#| message: false
#| warning: false

library(tidyverse) # includes and loads the packages listed below
# library(readr)
# library(tidyr)
# library(dplyr)
# library(ggplot2)
# library(lubridate)

library(stringr)  # for manipulating text strings

library(here) # helps with paths for files
library(arsenal)

library(purrr)

library(broom) # for creating tibbles of model output information

library(knitr) # for formatting tables
library(kableExtra) # for formatting tables

# libraries for spatial data & animating it...
library(maps)
library(ggmap)
library(gganimate)
library(gifski) # for gif output
library(av) # for video output
library(png)
library(tools)

library(sf)
library(geosphere)  # for calculating change in bearing and distances for gCOBs


# for maps with polygons for countries and states
library(rnaturalearth)
library(rnaturalearthdata)

# library(scales)
# library(ggbeeswarm)

```



### load custom functions
```{r source custom functions to simplify conducting these analyses}
#| echo: false

# getwd()

# source("R_scripts/functions/fish_calculate_gCOB.R")
# source("R_scripts/functions/filter_gCOB_dataframes.R")
# source("R_scripts/functions/calculate_gCOB_movement.R")
# source("R_scripts/functions/calculate_gCOB_annual_movement.R")
# source("R_scripts/functions/animate_fish_species_map.R")

# source("R_scripts/functions/test_gCOBs_over_time.R")
# source("R_scripts/functions/combine_and_modify_models.R")
# source("R_scripts/functions/plot_lat_long_gCOB_data.R")


```


### load data
```{r load data from csv file and filter for desired years and months}
#| echo: false
#| message: false

# import csv of clean bio data
bio_data_clean_climate <- read_csv(file = "data/clean_data/bio_data_clean_climate.csv")

# # import csv with climate zone information for each fish species as web scraped from FishBase
# species_climate_df <- read_csv(file = "output/species_climate_data.csv")


```



# IF USING DATA FROM .CSV FILES - MUST REDO GEOMETRIES

## Convert latitude and longitude into spatial data
 - Use coordinate reference system (crs) of *NAD83*, which is good for North America. It's specifically tied to the North American tectonic plate to minimize changes to coordinate values over time.
      - versus WGS84, which is global.
 - *NAD83 EPSG code = 4269*
 - WGS84 EPSG code = 4326
 - The underlying ellipsoid model is different for NAD83 versus WGS84.
      - NAD83 uses the GRS80 ellipsoid
      - WGS84 uses the WGS84 ellipsoid
- Ellipsoid parameter values
      - *NAD83*: 
          - semi-major axis ('a'): 6378137.0 meters
          - flattening ('f'): 1 / 298.257222101
      - WGS84: 
          - semi-major axis ('a'): 6378137.0 meters
          - flattening ('f'): 1 / 298.257223563



### convert to NAD83 coordinate reference system
```{r}

bio_data_clean_climate <- st_as_sf(bio_data_clean_climate, coords=c("DECSLON", "DECSLAT"), crs=4269, remove = FALSE)

```



### create base map of GoM
```{r obtain GoM map}
#| echo: false
#| message: false


# Get world polygons (already in sf format)
world_base_map_sf <- ne_countries(scale = "medium", returnclass = "sf")

# Get US state polygons (already in sf format)
usa_state_borders_sf <- ne_states(country = "United States of America", returnclass = "sf")

# # Check the CRS (default is 4326)
# st_crs(world_base_map_sf)  # Should output EPSG:4326
# st_crs(usa_state_borders_sf)  # Should output EPSG:4326

# Transform the CRS to 4269 (if needed)
world_base_map_sf <- st_transform(world_base_map_sf, 4269)
usa_state_borders_sf <- st_transform(usa_state_borders_sf, 4269)


# Plot the base map
base_map <- ggplot() +
  geom_sf(data = world_base_map_sf, color = "black", fill = "white") +
  geom_sf(data = usa_state_borders_sf, color = "black", fill = "white") +
  coord_sf(xlim = c(-97.7, -79.9), ylim = c(24, 31), crs = st_crs(4269)) +
  theme(panel.background = element_rect(fill = c("#CFEFFF")),
        panel.border = element_rect(fill = NA, color = "black")) +
  labs(x = "Longitude", y = "Latitude")

#plot base_map
base_map

# # Add point geometries to the base map
# map_with_points <- base_map +
#   geom_sf(data = bio_data_clean_yr1983_2022_m678_western, color = "red", size = 1.5)
# 
# # Plot the map with points
# map_with_points


```




## trawl stations per year - summarize data

```{r get number of trawl stations sampled each year}
#| echo: false

data_to_use_here_3 <- bio_data_clean_climate


# Summarize station data while preserving geometry
station_data <- data_to_use_here_3 %>%
  group_by(STATIONID) %>%
  summarize(CRUISEID = unique(CRUISEID),
            VESSEL = unique(VESSEL),
            CRUISE_NO = unique(CRUISE_NO),
            P_STA_NO = unique(P_STA_NO),
            YR = unique(YR),
            DEPTH_SSTA = unique(DEPTH_SSTA),
            MO_DAY_YR = unique(MO_DAY_YR),
            TEMP_SSURF = unique(TEMP_SSURF),
            TEMP_BOT = unique(TEMP_BOT),
            TEMP_SAIR = unique(TEMP_SAIR),
            geometry = st_geometry(first(geometry)))  # Retain geometry column


stations_num_yr <- station_data %>% 
  group_by(YR) %>% 
  summarize('# of trawls' = n(),
            months = paste(unique(month(dmy(MO_DAY_YR))), sep = ";", collapse = "; ")) %>% 
  rename(year = YR)


```




## animated map of trawl locations

```{r}

# Plot the base map with station points
station_maps <- ggplot() +
  geom_sf(data = world_base_map_sf, color = "black", fill = "white") +
  geom_sf(data = usa_state_borders_sf, color = "black", fill = "white") +
  geom_sf(data = station_data,
          aes(geometry = geometry, fill = YR),  # Use geometry for plotting
          size = 0.8,
          shape = 21,
          color = "maroon4",
          # fill = "darkorange4",
          stroke = 0.05) +    # width of border line
  coord_sf(xlim = c(-97.7, -79.9), ylim = c(24, 31), crs = st_crs(4269)) +
  theme(panel.background = element_rect(fill = c("#CFEFFF")),
        panel.border = element_rect(fill = NA, color = "black"),
        text = element_text(family = "Times", size = 12),
        plot.title = element_text(face = "bold", hjust = 0.5),
        axis.title = element_text(size = 12, face = "bold"),
        axis.text = element_text(size = 10),
        legend.title = element_text(size = 12, face = "bold", hjust = 0.5),
        legend.text = element_text(size = 10),
        legend.position = "none") +
  labs(title = "Trawl Stations for entire GoM\n1982 - 2022 all months",
       x = "Longitude", y = "Latitude")


# Preview the map
print(station_maps)


# Animate the map
animation <- station_maps +
    transition_time(YR) +
    ggtitle(label = NULL,  # No main title, just a subtitle
            subtitle = bquote(.(paste0("GoM 1982-2022\nYear: {frame_time}\nFrame {frame} of {nframes}")))) +
    shadow_mark(alpha = 0.6,
                color = "darkgray",
                size = 1.1) +
  scale_fill_gradient(low = "lightpink", high = "maroon4",
    limits = range(station_data$YR),
    breaks = seq(min(station_data$YR), max(station_data$YR), by = 2),
    labels = seq(min(station_data$YR), max(station_data$YR), by = 2),
    name = "Year")


num_years <- max(station_data$YR - min(station_data$YR)) + 1
animated_station_maps <- animate(animation,
                                 nframes = num_years,
                                 fps = 2,
                                 duration = (num_years + 5)/2,
                                 end_pause = 5,
                                 height = 4,
                                 width = 6,
                                 units = "in",
                                 res = 300)

animated_station_maps

anim_save("AFS_figures/allGoM_allYR_allmonths_allstations_animated_map1.gif", animation = animated_station_maps)


```







