---
title: "Analysis of SEAMAP data for the entire GoM for the years 2010 - 2022 and fall sampling months of October, November, and December."
author: "Shayna A. Sura"
format: html
---

# Tropicalization of Gulf of Mexico Fish Communities

## Analysis of SEAMAP Trawl Data

### Data Subset: all of GoM stations for fall trawls (October, November, December) in years 2010 - 2022

```{r code to render to pdf format}
#| echo: false
#| eval: false

quarto::quarto_render(input = "R_scripts/7_AFS_allGoM_fall.qmd", output_format = "pdf")

```



```{r load packages}
#| echo: false
#| message: false
#| warning: false

library(tidyverse) # includes and loads the packages listed below
# library(readr)
# library(tidyr)
# library(dplyr)
# library(ggplot2)
# library(lubridate)

library(stringr)  # for manipulating text strings

library(here) # helps with paths for files
library(arsenal)

library(purrr)

library(broom) # for creating tibbles of model output information

library(knitr) # for formatting tables
library(kableExtra) # for formatting tables

# libraries for spatial data & animating it...
library(maps)
library(ggmap)
library(gganimate)
library(gifski) # for gif output
library(av) # for video output
library(png)
library(tools)

library(sf)
library(geosphere)  # for calculating change in bearing and distances for gCOBs


# for maps with polygons for countries and states
library(rnaturalearth)
library(rnaturalearthdata)

# library(scales)
# library(ggbeeswarm)

```

```{r source custom functions to simplify conducting these analyses}
#| echo: false

# getwd()


source("R_scripts/functions/fish_calculate_gCOB.R")
source("R_scripts/functions/filter_gCOB_dataframes.R")
source("R_scripts/functions/calculate_gCOB_movement.R")
source("R_scripts/functions/calculate_gCOB_annual_movement.R")
source("R_scripts/functions/animate_fish_species_map.R")

# source("R_scripts/functions/test_gCOBs_over_time.R")
# source("R_scripts/functions/combine_and_modify_models.R")
# source("R_scripts/functions/plot_lat_long_gCOB_data.R")


```




```{r load data from csv file and filter for desired years and months}
#| echo: false
#| message: false

# import csv of clean bio data
bio_data_clean <- read_csv(file = "data/clean_data/bio_data_clean.csv")

# import csv with climate zone information for each fish species as web scraped from FishBase
species_climate_df <- read_csv(file = "output/species_climate_data.csv")


# subset the data for the target years and months
years_to_include <- c(2010:2022)
months_to_include <- c(10,11,12)


bio_data_clean_yr2010_2022_m101112 <- bio_data_clean %>% 
  filter(YR %in% years_to_include) %>% 
  mutate(MONTH = month(dmy(MO_DAY_YR))) %>% 
  filter(MONTH %in% months_to_include)


# export data subset to csv file
write_csv(bio_data_clean_yr2010_2022_m101112, file = "data/clean_data/bio_data_clean_yr2010_2022_m101112.csv")

```



# IF USING DATA FROM .CSV FILES - MUST REDO GEOMETRIES

## Convert latitude and longitude into spatial data
 - Use coordinate reference system (crs) of *NAD83*, which is good for North America. It's specifically tied to the North American tectonic plate to minimize changes to coordinate values over time.
      - versus WGS84, which is global.
 - *NAD83 EPSG code = 4269*
 - WGS84 EPSG code = 4326
 - The underlying ellipsoid model is different for NAD83 versus WGS84.
      - NAD83 uses the GRS80 ellipsoid
      - WGS84 uses the WGS84 ellipsoid
- Ellipsoid parameter values
      - *NAD83*: 
          - semi-major axis ('a'): 6378137.0 meters
          - flattening ('f'): 1 / 298.257222101
      - WGS84: 
          - semi-major axis ('a'): 6378137.0 meters
          - flattening ('f'): 1 / 298.257223563

```{r}

# bio_data_clean <- st_as_sf(bio_data_clean, coords=c("DECSLON", "DECSLAT"), crs=4269, remove = FALSE)

bio_data_clean_yr2010_2022_m101112 <- st_as_sf(bio_data_clean_yr2010_2022_m101112, coords=c("DECSLON", "DECSLAT"), crs=4269, remove = FALSE)

```






```{r obtain GoM map}
#| echo: false
#| message: false


# Get world polygons (already in sf format)
world_base_map_sf <- ne_countries(scale = "medium", returnclass = "sf")

# Get US state polygons (already in sf format)
usa_state_borders_sf <- ne_states(country = "United States of America", returnclass = "sf")

# Check the CRS (default is 4326)
st_crs(world_base_map_sf)  # Should output EPSG:4326
st_crs(usa_state_borders_sf)  # Should output EPSG:4326

# Transform the CRS to 4269 (if needed)
world_base_map_sf <- st_transform(world_base_map_sf, 4269)
usa_state_borders_sf <- st_transform(usa_state_borders_sf, 4269)


# Plot the base map
base_map <- ggplot() +
  geom_sf(data = world_base_map_sf, color = "black", fill = "white") +
  geom_sf(data = usa_state_borders_sf, color = "black", fill = "white") +
  coord_sf(xlim = c(-97.7, -79.9), ylim = c(24, 31), crs = st_crs(4269)) +
  theme(panel.background = element_rect(fill = c("#CFEFFF")),
        panel.border = element_rect(fill = NA, color = "black")) +
  labs(x = "Longitude", y = "Latitude")

#plot base_map
base_map

# # Add point geometries to the base map
# map_with_points <- base_map +
#   geom_sf(data = bio_data_clean_yr1983_2022_m678_western, color = "red", size = 1.5)
# 
# # Plot the map with points
# map_with_points


```






```{r}

## species list for 2010 - 2022 fall months (10,11,12) data only
species_df_yr2010_2022_m101112 <- bio_data_clean_yr2010_2022_m101112 %>% 
  select(new_TAXON, RANK) %>% 
  filter(RANK == "SPECIES") %>% 
  distinct(new_TAXON, .keep_all = TRUE) %>% 
  arrange(new_TAXON)


# dim(species_df_yr2010_2022_m101112) # 500 unique taxonomic names
species_vec_yr2010_2022_m101112 <- species_df_yr2010_2022_m101112$new_TAXON

```


```{r}

# Step 1 - use fish_calculate_gCOB function
# Applying the gCOB calculate function to each unique species for the dataset only containing 2010 - 2022 data
# create an empty list to store cg dataframe
gCOBs_yr2010_2022_m101112 <- list()

# Apply the fish_calculate_gCOB function to each unique species and combine results into a list
result_list_yr2010_2022_m101112 <- purrr::map(species_vec_yr2010_2022_m101112, ~fish_calculate_gCOB(data = bio_data_clean_yr2010_2022_m101112,
                                                                                                    output_name = "gCOBs_yr2010_2022_m101112",
                                                                                                    taxonomic_column = "new_TAXON",
                                                                                                    common_name_column = "new_TAXON",
                                                                                                    species_name = .))



# Step 2 - filter to only include fish species with 3+ gCOB observations
# Filter the list of data frames
gCOBs_yr2010_2022_m101112 <- filter_gCOB_dataframes(gCOBs_yr2010_2022_m101112, min_years = 3)

# get updated list of fish species names
species_vec_yr2010_2022_m101112 <- gsub("_gCOB", "", names(gCOBs_yr2010_2022_m101112))

# get number of fish species in filtered dataset
filtered_num_species <- length(species_vec_yr2010_2022_m101112)  #347 species with at least 3 years of data



# Step 3 - calculate change in gCOBs over time in terms of bearing and distance

#initiate empty list
movement_results <- list()

gCOBs_data <- get("gCOBs_yr2010_2022_m101112", envir = .GlobalEnv)
movement_results <- calculate_all_gCOB_movements(gCOBs_data)


# Step 3 part 2 - calculate change in gCOBs from year to year (bearing and distance) for each species

gCOBs_data <- get("gCOBs_yr2010_2022_m101112", envir = .GlobalEnv)
annual_movement_results <- calculate_annual_gCOB_movements(gCOBs_data)

```


## show gCOBs for a subset of species as examples

```{r}

lane_snapper_data <- bio_data_clean_yr2010_2022_m101112 %>% 
  filter(new_TAXON == "LUTJANUS SYNAGRIS") %>% 
  mutate(common_name = "Lane snapper")

lane_snapper_gCOBs <- as.data.frame(gCOBs_yr2010_2022_m101112["LUTJANUS SYNAGRIS_gCOB"])
lane_snapper_gCOBs <- lane_snapper_gCOBs %>% 
  rename(year = LUTJANUS.SYNAGRIS_gCOB.year,
         latitude = LUTJANUS.SYNAGRIS_gCOB.latitude,
         longitude = LUTJANUS.SYNAGRIS_gCOB.longitude)


lane_snapper_annual_movements <- annual_movement_results %>% 
  filter(species == "LUTJANUS SYNAGRIS")




# ## map of movement of gCOBs - NOT WORKING YET
# # sequence of labels that are not every 10 degrees
# less_bin_labels <- c("0", rep("",8), "90", rep("",8), "180", rep("",8), "270", rep("",8))
# 
# ggplot(data = lane_snapper_annual_movements, aes(x = bearing, y = distance)) +
#   geom_bar(stat = "identity", width = 1, color = "black") +
#   coord_polar(theta = "x", start = -pi/32, direction = 1) +   # Adjust start for correct orientation (due north at 12 o'clock)
#   scale_y_continuous(expand = c(0, 0)) +
#   # scale_x_continuous(limits = c(0,360),
#   #                  # drop = FALSE,
#   #                  # labels = less_bin_labels,    # less labels
#   #                  expand = c(0,0.5)
#   #                  ) +
#   ggtitle("Lane Snapper\nWind Rose Plot with Distance")
  

```


# animated map of lane snapper data
```{r}

# ## map of raw data points
# lane_snapper_map <- ggplot() +
#   geom_sf(data = world_base_map_sf, color = "black", fill = "white") +
#   geom_sf(data = usa_state_borders_sf, color = "black", fill = "white") +
#   geom_sf(data = lane_snapper_data, 
#           aes(geometry = geometry), size = 3) +
#   coord_sf(xlim = c(-97.7, -79.9), ylim = c(24, 31), crs = st_crs(4269)) +
#   theme(panel.background = element_rect(fill = c("#CFEFFF")),
#         panel.border = element_rect(fill = NA, color = "black")) +
#   labs(x = "Longitude", y = "Latitude")


## animated map of lane snapper SEAMAP data
lane_snapper_function_animation <- fish_species_animation(
  data = lane_snapper_data,
  taxonomic_column = "new_TAXON",
  common_name_column = "common_name",
  species_name = "LUTJANUS SYNAGRIS",
  low_color = "#FADFE4",
  high_color = "#DB1133"
)

lane_snapper_function_animation

anim_save("plots/lane_snapper_map1.gif")


```


## map of lane snapper annual gCOBs

```{r}

## map of gCOBs
lane_snapper_gCOBs_map <- base_map +
                            geom_point(data = lane_snapper_gCOBs, aes(x = longitude, y = latitude, fill = year),
                                       size = 5,
                                       stroke = 0.7,
                                       shape = 21,
                                       color = "black")  +
                            scale_fill_viridis_c(alpha = 0.7)

lane_snapper_gCOBs_map


# Save the combined plot
ggsave("AFS_figures/lane_snapper_gCOBs_map.png", lane_snapper_gCOBs_map, height = 6, width = 8)


# getwd()


# ggplot(lane_snapper_gCOBs, aes(longitude, latitude)) +
#   geom_point(aes(colour = year),
#                  size = 5
#              )
# 
# ggplot(lane_snapper_gCOBs, aes(year, longitude)) +
#   geom_point() +
#   geom_smooth(method = lm,
#               level = 0.95,
#               col = "purple",
#               linewidth = 2)
# 
# ggplot(lane_snapper_gCOBs, aes(year, latitude)) +
#   geom_point() +
#   geom_smooth(method = lm,
#               level = 0.95,
#               col = "purple",
#               linewidth = 2)

```


# plot of lane snapper bearing and distance changes in gCOBs over time
```{r}

lane_snapper_annual_movements

ls_avg_bearing = mean(lane_snapper_annual_movements$bearing)
ls_SD_bearing = sd(lane_snapper_annual_movements$bearing)

ls_avg_bearing
ls_SD_bearing

# Calculate mean, SD, and SE
summary_stats <- lane_snapper_annual_movements %>%
  summarise(
    mean_bearing = mean(bearing, na.rm = TRUE),
    sd_bearing = sd(bearing, na.rm = TRUE),
    se_bearing = sd_bearing / sqrt(n())
  ) %>% 
  mutate(num_species = c(1))


lane_snapper_annual_movements_fig1 <-
  ggplot(data = lane_snapper_annual_movements, aes(x = year_pair, y = bearing)) +
            geom_bar(stat = "identity", width = 1, color = "black", fill = "lightgrey") +
            geom_hline(yintercept = summary_stats$mean_bearing, color = "red", linetype = "solid", linewidth = 3, label = "Mean") +
            geom_hline(yintercept = summary_stats$mean_bearing + summary_stats$sd_bearing, color = "black", linetype = "dotted", linewidth = 1, label = "Mean + SD") +
            geom_hline(yintercept = summary_stats$mean_bearing - summary_stats$sd_bearing, color = "black", linetype = "dotted", linewidth = 1, label = "Mean - SD") +
            geom_hline(yintercept = summary_stats$mean_bearing + summary_stats$se_bearing, color = "green", linetype = "dashed", linewidth = 1, label = "Mean + SE") +
            geom_hline(yintercept = summary_stats$mean_bearing - summary_stats$se_bearing, color = "green", linetype = "dashed", linewidth = 1, label = "Mean - SE") +
            labs(x = "Year Pair", y = "Compass Bearing", title = "Lane Snapper gCOB Annual Movements") +
            theme_minimal() +
            theme(axis.text.x = element_text(angle = 45, hjust = 1),
                  panel.grid.major = element_blank(),  # Remove major grid lines
                  panel.grid.minor = element_blank()) +    # Remove minor grid lines
            scale_y_continuous(breaks = c(0, 90, 180, 270, 360))
  
# Save the combined plot
ggsave("AFS_figures/lane_snapper_gCOBs_annual_movements.png", lane_snapper_annual_movements_fig1, height = 3, width = 6)


```






## wind rose plots showing averages

```{r}

# 
# # Create a blank data frame with all compass directions
# # Assume a total of 36 bins (0-10, 10-20, ..., 350-360)
# angles <- seq(0, 350, by = 10)
# blank_data <- data.frame(
#   bearing = factor(angles, levels = angles), # Create factor to ensure all levels are displayed
#   num_species = rep(0, length(angles))       # Set to 0 as this is a blank plot
# )
# 
# # Extract labels for bins based on their midpoints
# bin_labels <- seq(0, 360, by = 10) # Midpoints of 36 bins (0-10, 10-20, ..., 350-360)
# 
# # Labels for key angles (0, 90, 180, 270)
# less_bin_labels <- c("0", rep("",8), "90", rep("",8), "180", rep("",8), "270", rep("",8))
# 
# # # Ensure the correct order of bearing bins
# # binned_climate_gCOBs <- binned_climate_gCOBs %>%
# #   mutate(bearing_bins = factor(bearing_bins,
# #                                levels = levels(bearing_bins),
# #                                labels = bin_labels[1:36]))
# 
# # Create the wind rose plot
# ggplot(data = summary_stats, aes(x = mean_bearing, y = num_species)) +
#   geom_bar(stat = "identity", width = 1, color = "black") +
#   geom_bar(data = summary_stats, stat = "identity", aes(x = mean_bearing, y = num_species),
#            width = 1, fill = "blue", alpha = 0.7) +
#   coord_polar(theta = "x", start = -pi/32, direction = 1) +   # Adjust start for correct orientation (due north at 12 o'clock)
#   scale_fill_manual(values = c("red")) +
#   scale_y_continuous(expand = c(0, 0)) +
#   scale_x_discrete(drop = FALSE,
#                    labels = less_bin_labels,    # Less labels for 0, 90, 180, 270 degrees
#                    expand = c(0,0.5)) +  
#   geom_text(aes(label = num_species), position = position_stack(vjust = 0.9), size = 3) +  # Add labels to petals
#   theme_minimal() +  # Apply a minimal theme for a clean look
#   # theme(axis.title = element_blank(),
#   #       axis.text.x = element_text(size = 12, vjust = 0.5),
#   #       axis.text.y = element_blank(),
#   #       panel.grid = element_blank(),
#   #       legend.position = "right") +  # Position the legend to the right
#   labs(fill = "Climate Distribution") +
#   ggtitle("Wind Rose Plot with Averaged gCOBs and Climate Distribution")


```



```{r}

library(ggplot2)
library(dplyr)

# Create a blank data frame with all compass directions
# Assume a total of 36 bins (0-10, 10-20, ..., 350-360)
angles <- seq(0, 350, by = 10)
blank_data <- data.frame(
  bearing = factor(angles, levels = angles), # Create factor to ensure all levels are displayed
  num_species = rep(0, length(angles))       # Set to 0 as this is a blank plot
)

# Labels for key angles (0, 90, 180, 270)
less_bin_labels <- c("0", rep("", 8), "90", rep("", 8), "180", rep("", 8), "270", rep("", 8))

# Create the blank wind rose plot
ggplot(data = blank_data, aes(x = bearing, y = num_species)) +
  geom_bar(stat = "identity", width = 1, color = "black", fill = "lightgrey") +
  # geom_bar(data = summary_stats, stat = "identity", aes(x = mean_bearing, y = num_species),
  #          width = 1, fill = "blue", alpha = 0.7) +
  coord_polar(theta = "x", start = -pi/36) +  # Adjust start to align 0 degrees at the top
  scale_y_continuous(expand = c(0, 0)) +
  scale_x_discrete(labels = less_bin_labels, expand = c(0,0.5)) +
  theme_minimal() +
  # theme(
  #   axis.title = element_blank(),
  #   axis.text.x = element_text(size = 12, hjust = 1),
  #   axis.text.y = element_blank(),
  #   panel.grid = element_blank(),
  #   panel.grid.major = element_blank(),
  #   panel.grid.minor = element_blank()
  # ) +
  ggtitle("Blank Wind Rose Plot with All Compass Directions")


```


```{r}

# Create a blank data frame with all compass directions
# Assume a total of 36 bins (0-10, 10-20, ..., 350-360)
angles <- as.numeric(seq(0, 360, by = 1))
angles <- as.numeric(c(seq(0, 359, by = 1), 359.999999999))
blank_data <- data.frame(
  bearing = (angles), # Create factor to ensure all levels are displayed
  num_species = as.numeric(rep(0, length(angles)))       # Set to 0 as this is a blank plot
)

# Labels for key angles (0, 90, 180, 270)
less_bin_labels <- c("0", rep("", 8), "90", rep("", 8), "180", rep("", 8), "270", rep("", 8))


# Create the blank wind rose plot
lane_snapper_windrose_Fig1 <- ggplot(data = blank_data, aes(x = bearing, y = num_species)) +
                                  geom_bar(stat = "identity", width = 1, color = "black", fill = "lightgrey") +
                                  geom_bar(data = summary_stats, stat = "identity", aes(x = mean_bearing, y = num_species),
                                           width = 8, fill = "red") +
                                  coord_polar(theta = "x", start = 0, direction = 1) +  # adjust start to align 0 degrees at the top
                                  scale_y_continuous(expand = c(0, 0),
                                                     limits = c(0,1)) +
                                  scale_x_continuous(breaks = c(0,90,180,270),
                                                     labels = c("0", "90", "180", "270"),
                                                     expand = c(0,0)) +
                                  theme_minimal() +
                                  theme(axis.text.x = element_text(size = 16, face = "bold",
                                                                   margin = margin(t = 5),  # Add margin to avoid overlap with outer axis line
                                                                   vjust = 0.5, hjust = 0.5),  
                                    axis.title.x = element_text(size = 12, face = "bold"),  # Bold x-axis label
                                    axis.title.y = element_blank(),
                                    axis.text.y = element_blank(),
                                    panel.grid.major.x = element_line(color = "gray", size = 1),
                                    panel.grid.minor.x = element_blank(),
                                    panel.grid.major.y = element_line(color = "gray", size = 0.5),
                                    panel.grid.minor.y = element_blank(),
                                    panel.border = element_blank()
                                  ) +
                                  labs(x = "Compass Bearing",
                                       y = "Number of Species")


# Save the combined plot
ggsave("AFS_figures/lane_snapper_windrose_Fig1.png", lane_snapper_windrose_Fig1, height = 6, width = 6)




# Create the blank wind rose plot
ggplot(data = blank_data, aes(x = bearing, y = num_species)) +
  geom_bar(stat = "identity", width = 1, color = "black", fill = "lightgrey") +
  geom_bar(data = summary_stats, stat = "identity", aes(x = mean_bearing, y = num_species),
           width = 8, fill = "red") +
  coord_polar(theta = "x", start = 0, direction = 1) +  # adjust start to align 0 degrees at the top
  scale_y_continuous(expand = c(0, 0),
                     limits = c(0,1)) +
  scale_x_continuous(breaks = c(0,90,180,270),
                     labels = c("0", "90", "180", "270"),
                     expand = c(0,0)) +
  theme_minimal() +
  theme(axis.text.x = element_text(size = 16, face = "bold",
                                   margin = margin(t = 5),  # Add margin to avoid overlap with outer axis line
                                   vjust = 0.5, hjust = 0.5),  
    axis.text.y = element_text(size = 12),  # Optional: Adjust size of y-axis labels if needed
    axis.title.x = element_text(size = 12, face = "bold"),  # Bold x-axis label
    axis.title.y = element_text(size = 12, face = "bold"),  # Bold y-axis label
    panel.grid.major.x = element_line(color = "gray", size = 1),
    panel.grid.minor.x = element_blank(),
    panel.grid.major.y = element_line(color = "gray", size = 0.5),
    panel.grid.minor.y = element_blank(),
    panel.border = element_blank()
    # axis.line.y.right = element_blank()
  ) +
  labs(x = "Compass Bearing",      # Set x-axis label
       y = "Number of Species",    # Set y-axis label
       title = "Blank Wind Rose Plot with All Compass Directions")
  # ggtitle("Blank Wind Rose Plot with All Compass Directions")

```



# wind rose plots for ALL FISH SPECIES - showing average shift in gCOBs over time

```{r}


# Function to calculate standard error
std_error <- function(x) {
  sd(x, na.rm = TRUE) / sqrt(sum(!is.na(x)))
}

# Summarize gCOB data by species: calculate mean, SD, and SE for bearing and distance
summarized_gCOBs <- annual_movement_results %>%
  group_by(species) %>%
  summarize(
    avg_bearing = mean(bearing, na.rm = TRUE),
    sd_bearing = sd(bearing, na.rm = TRUE),  # Standard deviation of bearing
    se_bearing = std_error(bearing),  # Standard error of bearing
    avg_distance = mean(distance, na.rm = TRUE),
    sd_distance = sd(distance, na.rm = TRUE),  # Standard deviation of distance
    se_distance = std_error(distance),  # Standard error of distance
    sample_size = n(),
    earliest_year = min(as.numeric(substring(year_pair, 1, 4)), na.rm = TRUE),
    latest_year = max(as.numeric(substring(year_pair, 6, 9)), na.rm = TRUE)
  ) %>%
  ungroup()



# Join with climate zone data
binned_climate_gCOBs <- summarized_gCOBs %>%
  rename(species_name = species) %>%
  left_join(species_climate_df, join_by(species_name)) %>% 
  rename(species = species_name) %>%
  filter(climate_zone != "deep-water") %>% 
  # Bin the average bearing into 36 bins
  mutate(bearing_bins = cut(avg_bearing, breaks = seq(0, 360, length.out = 37), include.lowest = TRUE)) %>%
  group_by(climate_zone, bearing_bins) %>%
  summarize(
    num_species = n(),
    avg_distance = mean(avg_distance, na.rm = TRUE),
    .groups = 'drop'
  ) %>%
  ungroup()

# change order of factor levels for climate zones
binned_climate_gCOBs$climate_zone <- factor(binned_climate_gCOBs$climate_zone, 
                                            levels = c("tropical", "subtropical", "temperate"))

# Verify the changes
levels(binned_climate_gCOBs$climate_zone)





# Extract labels for bins based on their midpoints
bin_labels <- seq(0, 360, by = 10) # Midpoints of 36 bins (0-10, 10-20, ..., 350-360)

# Labels for key angles (0, 90, 180, 270)
less_bin_labels <- c("0", rep("",8), "90", rep("",8), "180", rep("",8), "270", rep("",8))

# Ensure the correct order of bearing bins
binned_climate_gCOBs <- binned_climate_gCOBs %>%
  mutate(bearing_bins = factor(bearing_bins,
                               levels = levels(bearing_bins),
                               labels = bin_labels[1:36]))

# Create the wind rose plot
GoM_20102022_fall_windroses_fig1 <- 
  ggplot(data = binned_climate_gCOBs, aes(x = bearing_bins, y = num_species,
                                        fill = climate_zone,
                                        group = climate_zone)) +
      geom_bar(stat = "identity", width = 1, color = "black") +
      coord_polar(theta = "x", start = -pi/32, direction = 1) +   # Adjust start for correct orientation (due north at 12 o'clock)
      scale_fill_manual(values = c("tropical" = "coral", "temperate" = "skyblue", "subtropical" = "lightgreen", "deep-water" = "navy")) +
      scale_y_continuous(expand = c(0, 0)) +
      scale_x_discrete(drop = FALSE,
                       labels = less_bin_labels,    # Less labels for 0, 90, 180, 270 degrees
                       expand = c(0,0.5)) +  
      geom_text(aes(label = num_species), position = position_stack(vjust = 0.9), size = 3) +  # Add labels to petals
    theme_minimal() +
    theme(plot.title = element_text(size = 18, face = "bold", vjust = 0.5, hjust = 0.5),
      axis.text.x = element_text(size = 14, face = "bold",
                                   margin = margin(t = 5),  # Add margin to avoid overlap with outer axis line
                                   vjust = 0.5, hjust = 0.5),  
      axis.text.y = element_text(size = 12),  # Optional: Adjust size of y-axis labels if needed
      axis.title.x = element_text(size = 12, face = "bold"),  # Bold x-axis label
      axis.title.y = element_text(size = 12, face = "bold"),  # Bold y-axis label
      panel.grid.major.x = element_line(color = "gray", size = 0.3),
      panel.grid.minor.x = element_blank(),
      panel.grid.major.y = element_line(color = "gray", size = 0.2),
      panel.grid.minor.y = element_blank(),
      panel.border = element_blank(),
      legend.position = "bottom",
      legend.title = element_blank(),
      strip.text.x = element_text(size = 16, face = "bold", vjust = 0.5, hjust = 0.5)
      # axis.line.y.right = element_blank()
    ) +
  labs(x = "Compass Bearing",      # Set x-axis label
       y = "Number of Species",    # Set y-axis label
       title = "GoM 2010-2022 Fall") +
  facet_wrap(~climate_zone, scales = "free_y", nrow = 3)  # Facet by climate zone with free y-scales

GoM_20102022_fall_windroses_fig1

# Save the combined plot
ggsave("AFS_figures/GoM_20102022_fall_windroses_fig1.png", GoM_20102022_fall_windroses_fig1, height = 11, width = 6)


```







########################################### DID NOT DO ANYTHING PAST HERE YET FOR AFS ##################################################
########################################################################################################################################





### Number of trawl stations sampled each year

```{r get number of trawl stations sampled each year}
#| echo: false

data_to_use_here_3 <- bio_data_clean_yr2010_2022_m101112


station_data <- data_to_use_here_3 %>% 
  group_by(STATIONID) %>% 
  summarize(CRUISEID = unique(CRUISEID),
            VESSEL = unique(VESSEL),
            CRUISE_NO = unique(CRUISE_NO),
            P_STA_NO = unique(P_STA_NO),
            YR = unique(YR),
            DEPTH_SSTA = unique(DEPTH_SSTA),
            MO_DAY_YR = unique(MO_DAY_YR),
            DECSLAT = unique(DECSLAT),
            DECSLON = unique(DECSLON),
            DECELAT = unique(DECELAT),
            DECELON = unique(DECELON),
            TEMP_SSURF = unique(TEMP_SSURF),
            TEMP_BOT = unique(TEMP_BOT),
            TEMP_SAIR = unique(TEMP_SAIR))


stations_num_yr <- station_data %>% 
  group_by(YR) %>% 
  summarize('# of trawls' = n(),
            months = paste(unique(month(dmy(MO_DAY_YR))), sep = ";", collapse = "; ")) %>% 
  rename(year = YR)

stations_num_yr_formatted <- kable(stations_num_yr)

write_csv(stations_num_yr, file = "output/stations_num_yr_yr2010_2022_m101112.csv")

```

**Table 1.** Number of trawl stations for each year of data and the fall months during which sampling took place. `r stations_num_yr_formatted`

### Map of trawl stations

```{r map the trawl station locations, fig.width=8.5, fig.height=8.5}
#| echo: false


station_maps <- base_map +
  geom_point(data = station_data,
             na.rm = FALSE,
             aes(x = DECSLON,
                 y = DECSLAT,
                 group = NA,
                 fill = YR),
             size = 0.6,
             shape = 21,
             color = "darkorange4",
             fill = "orange",
             stroke = 0.05) +    # width of border line
  # scale_fill_viridis_d(name = "Year") +
    # scale_fill_gradient(low = "darkgray",
    #                    high = "springgreen",
    #                    limits = c(),
    #                    # breaks = c(),
    #                    # labels = c(),
    #                    name = "Year") +
  theme(text = element_text(family = "Times",
                            size = 14),
        plot.title = element_text(face = "bold",
                                  hjust = 0.5),
        axis.title = element_text(size = 12,
                                  face = "bold"),
        axis.text = element_text(size = 10),
        legend.title = element_text(size = 12,
                                    face = "bold",
                                    hjust = 0.5),
        legend.text = element_text(size = 10),
        legend.position = "none") +
  labs(title = "2010 - 2022 October, November, December Data") +
  facet_wrap(~YR)


station_maps

ggsave("plots/station_maps_yr2010_2022_m101112.pdf", plot = station_maps, width = 9, height = 8.5)


```

**Figure 1.** Trawl sampling locations for each year of data collection for this subset of data (2010-2022 fall sampling in months of October, November, and December).




```{r get list of fish species taxonomic names}
#| echo: false

## species list for 2010 - 2022 fall months (10,11,12) data only
species_df_yr2010_2022_m101112 <- bio_data_clean_yr2010_2022_m101112 %>% 
  select(new_TAXON, RANK) %>% 
  filter(RANK == "SPECIES") %>% 
  distinct(new_TAXON, .keep_all = TRUE) %>% 
  arrange(new_TAXON)


# dim(species_df_yr2010_2022_m101112) # 500 unique taxonomic names
species_vec_yr2010_2022_m101112 <- species_df_yr2010_2022_m101112$new_TAXON
num_species <- length(species_vec_yr2010_2022_m101112) # 500 unique taxonomic names


```

There are `r num_species` fish species that were caught in trawls in this subset of data.

```{r gCOB analyses}
#| echo: false

# Step 1 - use fish_calculate_gCOB function
# Applying the gCOB calculate function to each unique species for the dataset only containing 2010 - 2022 data
# create an empty list to store cg dataframe
gCOBs_yr2010_2022_m101112 <- list()

# Apply the fish_calculate_gCOB function to each unique species and combine results into a list
result_list_yr2010_2022_m101112 <- purrr::map(species_vec_yr2010_2022_m101112, ~fish_calculate_gCOB(data = bio_data_clean_yr2010_2022_m101112,
                                                                                                    output_name = "gCOBs_yr2010_2022_m101112",
                                                                                                    taxonomic_column = "new_TAXON",
                                                                                                    common_name_column = "new_TAXON",
                                                                                                    species_name = .))



# Step 2 - filter to only include fish species with 5+ gCOB observations
# Filter the list of data frames
gCOBs_yr2010_2022_m101112 <- filter_gCOB_dataframes(gCOBs_yr2010_2022_m101112, min_years = 5)

# get updated list of fish species names
species_vec_yr2010_2022_m101112 <- gsub("_gCOB", "", names(gCOBs_yr2010_2022_m101112))

# get number of fish species in filtered dataset
filtered_num_species <- length(species_vec_yr2010_2022_m101112)



# Step 3 - use test_gCOBs_over_time function
# initiate empty data frame to store data in
gCOB_trends_yr2010_2022_m101112 <- data.frame(species = NA,
                                              latitude_year_coef = NA,
                                              latitude_year_SE = NA,
                                              latitude_pvalue = NA,
                                              latitude_rsquared = NA,
                                              latitude_rsquared_adj = NA,
                                              longitude_year_coef = NA,
                                              longitude_year_SE = NA,
                                              longitude_pvalue = NA,
                                              longitude_rsquared = NA,
                                              longitude_rsquared_adj = NA)

# initiate empty lists to store latitude and longitude model outputs
lat_models_yr2010_2022_m101112 <- list()
long_models_yr2010_2022_m101112 <- list()

# apply function
trends_results_yr2010_2022_m101112 <- purrr::map(species_vec_yr2010_2022_m101112,
                                                 ~ test_gCOBs_over_time(species_name = .,
                                                                        input_gCOBs = gCOBs_yr2010_2022_m101112,
                                                                        output_df_name = "gCOB_trends_yr2010_2022_m101112",
                                                                        latitude_models_output_name = "lat_models_yr2010_2022_m101112",
                                                                        longitude_models_output_name = "long_models_yr2010_2022_m101112"))

# remove row of NAs at beginning of dataframe
gCOB_trends_yr2010_2022_m101112 <- gCOB_trends_yr2010_2022_m101112 %>% 
  slice(-1)


# export results as csv file
write_csv(gCOB_trends_yr2010_2022_m101112, file = "output/gCOB_trends_results_raw_yr2010_2022_m101112.csv")


```

Of those `r num_species` fish species, only `r filtered_num_species` of them had gCOB values for at least 5 years. For these fish species, linear models were constructed to test if their latitude and longitude gCOBs had shifted over time.

```{r summarizing trends in gCOBs - pvalue summary}
#| echo: false

# Version 1 - summarizing based on p-values
# specify the gCOB trends dataframe to use here
gCOB_trends_data <- gCOB_trends_yr2010_2022_m101112 

# join climate zone information to gCOB_trends
climate_gCOB_trends <- gCOB_trends_data %>%
  rename(species_name = species) %>%
  left_join(species_climate_df, join_by(species_name)) %>% 
  rename(species = species_name)


# summarize the number of fish species that show significant shifts in latitude or longitude based on significant p-values
significant_pvalues <- climate_gCOB_trends %>% 
  group_by(climate_zone) %>% 
  summarize(n = n(),
            num_sig_pvalues_lat = length(which(latitude_pvalue < 0.05)),
            num_sig_pvalues_long = length(which(longitude_pvalue < 0.05)),
            num_sig_lat_shift_north = length(which(latitude_pvalue < 0.05 & latitude_year_coef > 0)),
            num_sig_lat_shift_south = length(which(latitude_pvalue < 0.05 & latitude_year_coef < 0)),
            num_sig_long_shift_west = length(which(longitude_pvalue < 0.05 & longitude_year_coef < 0)),
            num_sig_long_shift_east = length(which(longitude_pvalue < 0.05 & longitude_year_coef > 0))
  )
         
# significant_pvalues

# export to csv file
write_csv(significant_pvalues, file = "output/gCOB_trends_pvalue_summary_yr2010_2022_m101112.csv")


```

```{r summarizing trends in gCOBs - 95CI summary}
#| echo: false

# Step 1 - use combine_and_modify_models function to get single dataframe of model results from lists of model output
# latitudinal models
lat_models_df_yr2010_2022_m101112 <- combine_and_modify_models(model_list = lat_models_yr2010_2022_m101112,
                                                               output_name = "lat_models_df_yr2010_2022_m101112",
                                                               pattern = "latitude")
# longitudinal models
long_models_df_yr2010_2022_m101112 <- combine_and_modify_models(model_list = long_models_yr2010_2022_m101112,
                                                                output_name = "long_models_df_yr2010_2022_m101112",
                                                                pattern = "longitude")


# Step 2 - summarize trends based on 95CIs not overlapping with zero

# summarize the number of fish species that show significant shifts in latitude or longitude based on 95CIs that do not overlap with zero
## latitude models
significant_95CIs_lat <- lat_models_df_yr2010_2022_m101112 %>% 
  filter(term != "(Intercept)") %>%
  mutate(signif_95CI = ifelse(conf.low < 0 & conf.high > 0, "not significant", "significant")) %>% 
  mutate(signif_pvalue = ifelse(p.value < 0.05, "significant", "not significant")) %>% 
  group_by(climate_zone) %>% 
  summarize(n = n(),
            num_sig_95CIs_lat = sum(signif_95CI == "significant"),
            num_sig_pvalues_lat = sum(signif_pvalue == "significant"),
            num_sig_95CIs_lat_shift_north = sum(signif_95CI == "significant" & estimate > 0),
            num_sig_95CIs_lat_shift_south = sum(signif_95CI == "significant" & estimate < 0),
            north_shift_species = paste(species[signif_95CI == "significant" & estimate > 0], collapse = ", "),
            south_shift_species = paste(species[signif_95CI == "significant" & estimate < 0], collapse = ", ")
  )


## longitude models
significant_95CIs_long <- long_models_df_yr2010_2022_m101112 %>% 
  filter(term != "(Intercept)") %>%
  mutate(signif_95CI = ifelse(conf.low < 0 & conf.high > 0, "not significant", "significant")) %>% 
  mutate(signif_pvalue = ifelse(p.value < 0.05, "significant", "not significant")) %>% 
  group_by(climate_zone) %>% 
  summarize(n = n(),
            num_sig_95CIs_long = sum(signif_95CI == "significant"),
            num_sig_pvalues_long = sum(signif_pvalue == "significant"),
            num_sig_95CIs_long_shift_west = sum(signif_95CI == "significant" & estimate < 0),
            num_sig_95CIs_long_shift_east = sum(signif_95CI == "significant" & estimate > 0),
            west_shift_species = paste(species[signif_95CI == "significant" & estimate < 0], collapse = ", "),
            east_shift_species = paste(species[signif_95CI == "significant" & estimate > 0], collapse = ", ")
  )


## combine output from latitude and longitude summary dataframes
all_significant_95CIs <- significant_95CIs_lat %>% 
  left_join(significant_95CIs_long, join_by(climate_zone))


## export to csv file
write_csv(all_significant_95CIs, "output/gCOB_significant_trends_yr2010_2022_m101112.csv")


```





```{r formatting summary tables}
#| echo: false


table_2 <- all_significant_95CIs %>% 
  select(-north_shift_species, -south_shift_species, -west_shift_species, -east_shift_species,
         -n.y,
         -num_sig_pvalues_lat, -num_sig_pvalues_long) %>% 
    rename('n' = n.x,
         'significant latitude shift' = num_sig_95CIs_lat,
         'North shifted' = num_sig_95CIs_lat_shift_north,
         'South shifted'= num_sig_95CIs_lat_shift_south,
         'significant longitude shift' = num_sig_95CIs_long,
         'East shifted' = num_sig_95CIs_long_shift_east,
         'West shifted'= num_sig_95CIs_long_shift_west)

table_2_formatted <- knitr::kable(table_2)



```

**Table 2.** The number of fish species that have significantly shifted latitudinally or longitudinally based upon their gCOBs. `r table_2_formatted`





```{r getting separate tables for each climate zone with the species exhibiting each type of shift}
#| echo: false


species_tables <- all_significant_95CIs %>% 
  select(climate_zone, north_shift_species, south_shift_species, west_shift_species, east_shift_species)


# Modified code to ensure each species has a unique row identifier
extracted_data <- species_tables %>%
  pivot_longer(cols = -climate_zone, names_to = "direction", values_to = "species_list") %>%
  separate_rows(species_list, sep = ",\\s*") %>%
  filter(species_list != "") %>%
  mutate(direction = str_replace(direction, "_shift_species", "")) %>%
  group_by(climate_zone, direction) %>%
  mutate(row_id = row_number()) %>%
  ungroup()

# Get available direction column names
available_directions <- intersect(names(extracted_data), c("north", "south", "east", "west"))


# Pivot wider while keeping the unique row identifier
wide_data <- extracted_data %>%
  pivot_wider(names_from = direction, values_from = species_list, values_fn = list) %>%
  # glimpse()
  # unnest(c(south, west, east, north), names_sep = "_") %>%
  unnest(cols = all_of(available_directions), names_sep = "_") %>% 
  as.data.frame() %>% 
  # # replace NULL values with NA
  # mutate(across(all_of(available_directions), ~ ifelse(. == "NULL", NA, .))) %>% 
  select(-row_id) %>%
  arrange(climate_zone)


# Filter for the tropical climate zone
tropical_table <- wide_data %>%
  filter(climate_zone == "tropical") %>%
  select(-climate_zone)

# get directions of significant shifts
tropical_directions <- intersect(names(tropical_table), c("north", "south", "east", "west"))

# Output the tropical table
# tropical_table

table_3 <- tropical_table %>% 
  mutate(across(all_of(tropical_directions), str_to_sentence))

table_3_formatted <- knitr::kable(table_3)


# Filter for the subtropical climate zone
subtropical_table <- wide_data %>%
  filter(climate_zone == "subtropical") %>%
  select(-climate_zone)

# get directions of significant shifts
subtropical_directions <- intersect(names(subtropical_table), c("north", "south", "east", "west"))

table_4 <- subtropical_table %>% 
  mutate(across(all_of(subtropical_directions), str_to_sentence))

table_4_formatted <- knitr::kable(table_4)



# Filter for the temperate climate zone
temperate_table <- wide_data %>%
  filter(climate_zone == "temperate") %>%
  select(-climate_zone)

# get directions of significant shifts
temperate_directions <- intersect(names(temperate_table), c("north", "south", "east", "west"))

table_5 <- temperate_table %>% 
  mutate(across(all_of(temperate_directions), str_to_sentence))

table_5_formatted <- knitr::kable(table_5)


```

**Table XX.** The tropical fish species that have significantly shifted their geographical center of biomass (gCOB) over time.
`r table_3_formatted`

**Table XX.** The subtropical fish species that have significantly shifted their geographical center of biomass (gCOB) over time.
`r table_4_formatted`

**Table XX.** The temperate fish species that have significantly shifted their geographical center of biomass (gCOB) over time.
`r table_5_formatted`




```{r make figures of gCOB coefficient estimates}
#| echo: false

plot_lat_long_gCOB_data(lat_data = lat_models_df_yr2010_2022_m101112,
                        long_data = long_models_df_yr2010_2022_m101112,
                        title_prefix = "2010-2022 October, November, & December")


# should produce four figures
# Fig_1_yr2010_2022_m101112
# Fig_2_yr2010_2022_m101112
# Fig_3_yr2010_2022_m101112
# Fig_4_yr2010_2022_m101112


```


```{r Fig_2,  fig.width=8.5, fig.height=13}
#| echo: false

Fig_1_yr2010_2022_m101112

# fig.cap = "all species latitude gCOBs",

```

**Figure 2.** Latitudinal trends in gCOBs for fish species (deep-water, subtropical, temperate, tropical).




```{r Fig_3,  fig.width=8.5, fig.height=13}
#| echo: false

Fig_2_yr2010_2022_m101112

# fig.cap = "all species longitude gCOBs",

```

**Figure 3.**Longitudinal trends in gCOBs for fish species (deep-water, subtropical, temperate, tropical).



```{r Fig_4,  fig.width=8.5, fig.height=6.5}
#| echo: false

Fig_3_yr2010_2022_m101112

# fig.cap = "significant latitude gCOBs",

```

**Figure 4.**Latitudinal trends in gCOBs for fish species (deep-water, subtropical, temperate, tropical) with a significant shift over time.



```{r Fig_5,  fig.width=8.5, fig.height=5.5}
#| echo: false

Fig_4_yr2010_2022_m101112

# fig.cap = "significant longitude gCOBs",

```

**Figure 5.**Longitudinal trends in gCOBs for fish species (deep-water, subtropical, temperate, tropical) with a significant shift over time.




