---
title: "Analysis of SEAMAP data for the entire GoM for the years 2010 - 2022 and fall sampling months of October, November, and December."
author: "Shayna A. Sura"
format: html
---

# Tropicalization of Gulf of Mexico Fish Communities

## Analysis of SEAMAP Trawl Data

### Data Subset: all of GoM stations for fall trawls (October, November, December) in years 2010 - 2022

```{r code to render to pdf format}
#| echo: false
#| eval: false

quarto::quarto_render(input = "R_scripts/7_allGoM_fall.qmd", output_format = "pdf")

```



```{r load packages}
#| echo: false
#| message: false
#| warning: false

library(tidyverse) # includes and loads the packages listed below
# library(readr)
# library(tidyr)
# library(dplyr)
# library(ggplot2)
# library(lubridate)

library(stringr)  # for manipulating text strings

library(here) # helps with paths for files
library(arsenal)

library(purrr)

library(broom) # for creating tibbles of model output information

library(knitr) # for formatting tables
library(kableExtra) # for formatting tables

# libraries for spatial data & animating it...
library(maps)
library(ggmap)
library(gganimate)
library(gifski) # for gif output
library(av) # for video output
library(png)
library(tools)

# library(scales)
# library(ggbeeswarm)

```

```{r source custom functions to simplify conducting these analyses}
#| echo: false

# getwd()

source("R_scripts/functions/fish_calculate_gCOB.R")
source("R_scripts/functions/test_gCOBs_over_time.R")
source("R_scripts/functions/combine_and_modify_models.R")
source("R_scripts/functions/filter_gCOB_dataframes.R")
source("R_scripts/functions/plot_lat_long_gCOB_data.R")

```

```{r load data from csv file and filter for desired years and months}
#| echo: false
#| message: false

# import csv of clean bio data
bio_data_clean <- read_csv(file = "data/clean_data/bio_data_clean.csv")

# import csv with cimate zone information for each fish species as web scraped from FishBase
species_climate_df <- read_csv(file = "output/species_climate_data.csv")


# subset the data for the target years and months
years_to_include <- c(2010:2022)
months_to_include <- c(10,11,12)


bio_data_clean_yr2010_2022_m101112 <- bio_data_clean %>% 
  filter(YR %in% years_to_include) %>% 
  mutate(MONTH = month(dmy(MO_DAY_YR))) %>% 
  filter(MONTH %in% months_to_include)


# export data subset to csv file
write_csv(bio_data_clean_yr2010_2022_m101112, file = "data/clean_data/bio_data_clean_yr2010_2022_m101112.csv")

```


```{r obtain GoM map}
#| echo: false
#| message: false


gulf_of_mexico_info <- map_data("usa", region = ".")
world_base_map <- map_data("world", region = ".")
usa_state_borders <- map_data("state", region = ".")

base_map <- ggplot(data = world_base_map,
                   mapping = aes (x = long, y = lat, group = group)) +
  geom_polygon(color = "black",
               fill = "white") +
  geom_polygon(data = usa_state_borders,
               color = "black",
               fill = "white") + 
  coord_quickmap(xlim = c(-97.7, -79.9),
                 ylim = c(24, 31)) +
  theme(panel.background = element_rect(fill = c("#CFEFFF")),
        panel.border = element_rect(fill = NA,
                                    color = "black")
        ) +
  labs(x = "Longitude",
       y = "Latitude")

```

### Number of trawl stations sampled each year

```{r get number of trawl stations sampled each year}
#| echo: false

data_to_use_here_3 <- bio_data_clean_yr2010_2022_m101112


station_data <- data_to_use_here_3 %>% 
  group_by(STATIONID) %>% 
  summarize(CRUISEID = unique(CRUISEID),
            VESSEL = unique(VESSEL),
            CRUISE_NO = unique(CRUISE_NO),
            P_STA_NO = unique(P_STA_NO),
            YR = unique(YR),
            DEPTH_SSTA = unique(DEPTH_SSTA),
            MO_DAY_YR = unique(MO_DAY_YR),
            DECSLAT = unique(DECSLAT),
            DECSLON = unique(DECSLON),
            DECELAT = unique(DECELAT),
            DECELON = unique(DECELON),
            TEMP_SSURF = unique(TEMP_SSURF),
            TEMP_BOT = unique(TEMP_BOT),
            TEMP_SAIR = unique(TEMP_SAIR))


stations_num_yr <- station_data %>% 
  group_by(YR) %>% 
  summarize('# of trawls' = n(),
            months = paste(unique(month(dmy(MO_DAY_YR))), sep = ";", collapse = "; ")) %>% 
  rename(year = YR)

stations_num_yr_formatted <- kable(stations_num_yr)

write_csv(stations_num_yr, file = "output/stations_num_yr_yr2010_2022_m101112.csv")

```

**Table 1.** Number of trawl stations for each year of data and the fall months during which sampling took place. `r stations_num_yr_formatted`

### Map of trawl stations

```{r map the trawl station locations, fig.width=8.5, fig.height=8.5}
#| echo: false


station_maps <- base_map +
  geom_point(data = station_data,
             na.rm = FALSE,
             aes(x = DECSLON,
                 y = DECSLAT,
                 group = NA,
                 fill = YR),
             size = 0.6,
             shape = 21,
             color = "darkorange4",
             fill = "orange",
             stroke = 0.05) +    # width of border line
  # scale_fill_viridis_d(name = "Year") +
    # scale_fill_gradient(low = "darkgray",
    #                    high = "springgreen",
    #                    limits = c(),
    #                    # breaks = c(),
    #                    # labels = c(),
    #                    name = "Year") +
  theme(text = element_text(family = "Times",
                            size = 14),
        plot.title = element_text(face = "bold",
                                  hjust = 0.5),
        axis.title = element_text(size = 12,
                                  face = "bold"),
        axis.text = element_text(size = 10),
        legend.title = element_text(size = 12,
                                    face = "bold",
                                    hjust = 0.5),
        legend.text = element_text(size = 10),
        legend.position = "none") +
  labs(title = "2010 - 2022 October, November, December Data") +
  facet_wrap(~YR)


station_maps

ggsave("plots/station_maps_yr2010_2022_m101112.pdf", plot = station_maps, width = 9, height = 8.5)


```

**Figure 1.** Trawl sampling locations for each year of data collection for this subset of data (2010-2022 fall sampling in months of October, November, and December).




```{r get list of fish species taxonomic names}
#| echo: false

## species list for 2010 - 2022 fall months (10,11,12) data only
species_df_yr2010_2022_m101112 <- bio_data_clean_yr2010_2022_m101112 %>% 
  select(new_TAXON, RANK) %>% 
  filter(RANK == "SPECIES") %>% 
  distinct(new_TAXON, .keep_all = TRUE) %>% 
  arrange(new_TAXON)


# dim(species_df_yr2010_2022_m101112) # 500 unique taxonomic names
species_vec_yr2010_2022_m101112 <- species_df_yr2010_2022_m101112$new_TAXON
num_species <- length(species_vec_yr2010_2022_m101112) # 500 unique taxonomic names


```

There are `r num_species` fish species that were caught in trawls in this subset of data.

```{r gCOB analyses}
#| echo: false

# Step 1 - use fish_calculate_gCOB function
# Applying the gCOB calculate function to each unique species for the dataset only containing 2010 - 2022 data
# create an empty list to store cg dataframe
gCOBs_yr2010_2022_m101112 <- list()

# Apply the fish_calculate_gCOB function to each unique species and combine results into a list
result_list_yr2010_2022_m101112 <- purrr::map(species_vec_yr2010_2022_m101112, ~fish_calculate_gCOB(data = bio_data_clean_yr2010_2022_m101112,
                                                                                                    output_name = "gCOBs_yr2010_2022_m101112",
                                                                                                    taxonomic_column = "new_TAXON",
                                                                                                    common_name_column = "new_TAXON",
                                                                                                    species_name = .))



# Step 2 - filter to only include fish species with 5+ gCOB observations
# Filter the list of data frames
gCOBs_yr2010_2022_m101112 <- filter_gCOB_dataframes(gCOBs_yr2010_2022_m101112, min_years = 5)

# get updated list of fish species names
species_vec_yr2010_2022_m101112 <- gsub("_gCOB", "", names(gCOBs_yr2010_2022_m101112))

# get number of fish species in filtered dataset
filtered_num_species <- length(species_vec_yr2010_2022_m101112)



# Step 3 - use test_gCOBs_over_time function
# initiate empty data frame to store data in
gCOB_trends_yr2010_2022_m101112 <- data.frame(species = NA,
                                              latitude_year_coef = NA,
                                              latitude_year_SE = NA,
                                              latitude_pvalue = NA,
                                              latitude_rsquared = NA,
                                              latitude_rsquared_adj = NA,
                                              longitude_year_coef = NA,
                                              longitude_year_SE = NA,
                                              longitude_pvalue = NA,
                                              longitude_rsquared = NA,
                                              longitude_rsquared_adj = NA)

# initiate empty lists to store latitude and longitude model outputs
lat_models_yr2010_2022_m101112 <- list()
long_models_yr2010_2022_m101112 <- list()

# apply function
trends_results_yr2010_2022_m101112 <- purrr::map(species_vec_yr2010_2022_m101112,
                                                 ~ test_gCOBs_over_time(species_name = .,
                                                                        input_gCOBs = gCOBs_yr2010_2022_m101112,
                                                                        output_df_name = "gCOB_trends_yr2010_2022_m101112",
                                                                        latitude_models_output_name = "lat_models_yr2010_2022_m101112",
                                                                        longitude_models_output_name = "long_models_yr2010_2022_m101112"))

# remove row of NAs at beginning of dataframe
gCOB_trends_yr2010_2022_m101112 <- gCOB_trends_yr2010_2022_m101112 %>% 
  slice(-1)


# export results as csv file
write_csv(gCOB_trends_yr2010_2022_m101112, file = "output/gCOB_trends_results_raw_yr2010_2022_m101112.csv")


```

Of those `r num_species` fish species, only `r filtered_num_species` of them had gCOB values for at least 5 years. For these fish species, linear models were constructed to test if their latitude and longitude gCOBs had shifted over time.

```{r summarizing trends in gCOBs - pvalue summary}
#| echo: false

# Version 1 - summarizing based on p-values
# specify the gCOB trends dataframe to use here
gCOB_trends_data <- gCOB_trends_yr2010_2022_m101112 

# join climate zone information to gCOB_trends
climate_gCOB_trends <- gCOB_trends_data %>%
  rename(species_name = species) %>%
  left_join(species_climate_df, join_by(species_name)) %>% 
  rename(species = species_name)


# summarize the number of fish species that show significant shifts in latitude or longitude based on significant p-values
significant_pvalues <- climate_gCOB_trends %>% 
  group_by(climate_zone) %>% 
  summarize(n = n(),
            num_sig_pvalues_lat = length(which(latitude_pvalue < 0.05)),
            num_sig_pvalues_long = length(which(longitude_pvalue < 0.05)),
            num_sig_lat_shift_north = length(which(latitude_pvalue < 0.05 & latitude_year_coef > 0)),
            num_sig_lat_shift_south = length(which(latitude_pvalue < 0.05 & latitude_year_coef < 0)),
            num_sig_long_shift_west = length(which(longitude_pvalue < 0.05 & longitude_year_coef < 0)),
            num_sig_long_shift_east = length(which(longitude_pvalue < 0.05 & longitude_year_coef > 0))
  )
         
# significant_pvalues

# export to csv file
write_csv(significant_pvalues, file = "output/gCOB_trends_pvalue_summary_yr2010_2022_m101112.csv")


```

```{r summarizing trends in gCOBs - 95CI summary}
#| echo: false

# Step 1 - use combine_and_modify_models function to get single dataframe of model results from lists of model output
# latitudinal models
lat_models_df_yr2010_2022_m101112 <- combine_and_modify_models(model_list = lat_models_yr2010_2022_m101112,
                                                               output_name = "lat_models_df_yr2010_2022_m101112",
                                                               pattern = "latitude")
# longitudinal models
long_models_df_yr2010_2022_m101112 <- combine_and_modify_models(model_list = long_models_yr2010_2022_m101112,
                                                                output_name = "long_models_df_yr2010_2022_m101112",
                                                                pattern = "longitude")


# Step 2 - summarize trends based on 95CIs not overlapping with zero

# summarize the number of fish species that show significant shifts in latitude or longitude based on 95CIs that do not overlap with zero
## latitude models
significant_95CIs_lat <- lat_models_df_yr2010_2022_m101112 %>% 
  filter(term != "(Intercept)") %>%
  mutate(signif_95CI = ifelse(conf.low < 0 & conf.high > 0, "not significant", "significant")) %>% 
  mutate(signif_pvalue = ifelse(p.value < 0.05, "significant", "not significant")) %>% 
  group_by(climate_zone) %>% 
  summarize(n = n(),
            num_sig_95CIs_lat = sum(signif_95CI == "significant"),
            num_sig_pvalues_lat = sum(signif_pvalue == "significant"),
            num_sig_95CIs_lat_shift_north = sum(signif_95CI == "significant" & estimate > 0),
            num_sig_95CIs_lat_shift_south = sum(signif_95CI == "significant" & estimate < 0),
            north_shift_species = paste(species[signif_95CI == "significant" & estimate > 0], collapse = ", "),
            south_shift_species = paste(species[signif_95CI == "significant" & estimate < 0], collapse = ", ")
  )


## longitude models
significant_95CIs_long <- long_models_df_yr2010_2022_m101112 %>% 
  filter(term != "(Intercept)") %>%
  mutate(signif_95CI = ifelse(conf.low < 0 & conf.high > 0, "not significant", "significant")) %>% 
  mutate(signif_pvalue = ifelse(p.value < 0.05, "significant", "not significant")) %>% 
  group_by(climate_zone) %>% 
  summarize(n = n(),
            num_sig_95CIs_long = sum(signif_95CI == "significant"),
            num_sig_pvalues_long = sum(signif_pvalue == "significant"),
            num_sig_95CIs_long_shift_west = sum(signif_95CI == "significant" & estimate < 0),
            num_sig_95CIs_long_shift_east = sum(signif_95CI == "significant" & estimate > 0),
            west_shift_species = paste(species[signif_95CI == "significant" & estimate < 0], collapse = ", "),
            east_shift_species = paste(species[signif_95CI == "significant" & estimate > 0], collapse = ", ")
  )


## combine output from latitude and longitude summary dataframes
all_significant_95CIs <- significant_95CIs_lat %>% 
  left_join(significant_95CIs_long, join_by(climate_zone))


## export to csv file
write_csv(all_significant_95CIs, "output/gCOB_significant_trends_yr2010_2022_m101112.csv")


```





```{r formatting summary tables}
#| echo: false


table_2 <- all_significant_95CIs %>% 
  select(-north_shift_species, -south_shift_species, -west_shift_species, -east_shift_species,
         -n.y,
         -num_sig_pvalues_lat, -num_sig_pvalues_long) %>% 
    rename('n' = n.x,
         'significant latitude shift' = num_sig_95CIs_lat,
         'North shifted' = num_sig_95CIs_lat_shift_north,
         'South shifted'= num_sig_95CIs_lat_shift_south,
         'significant longitude shift' = num_sig_95CIs_long,
         'East shifted' = num_sig_95CIs_long_shift_east,
         'West shifted'= num_sig_95CIs_long_shift_west)

table_2_formatted <- knitr::kable(table_2)



```

**Table 2.** The number of fish species that have significantly shifted latitudinally or longitudinally based upon their gCOBs. `r table_2_formatted`





```{r getting separate tables for each climate zone with the species exhibiting each type of shift}
#| echo: false


species_tables <- all_significant_95CIs %>% 
  select(climate_zone, north_shift_species, south_shift_species, west_shift_species, east_shift_species)


# Modified code to ensure each species has a unique row identifier
extracted_data <- species_tables %>%
  pivot_longer(cols = -climate_zone, names_to = "direction", values_to = "species_list") %>%
  separate_rows(species_list, sep = ",\\s*") %>%
  filter(species_list != "") %>%
  mutate(direction = str_replace(direction, "_shift_species", "")) %>%
  group_by(climate_zone, direction) %>%
  mutate(row_id = row_number()) %>%
  ungroup()

# Get available direction column names
available_directions <- intersect(names(extracted_data), c("north", "south", "east", "west"))


# Pivot wider while keeping the unique row identifier
wide_data <- extracted_data %>%
  pivot_wider(names_from = direction, values_from = species_list, values_fn = list) %>%
  # glimpse()
  # unnest(c(south, west, east, north), names_sep = "_") %>%
  unnest(cols = all_of(available_directions), names_sep = "_") %>% 
  as.data.frame() %>% 
  # # replace NULL values with NA
  # mutate(across(all_of(available_directions), ~ ifelse(. == "NULL", NA, .))) %>% 
  select(-row_id) %>%
  arrange(climate_zone)


# Filter for the tropical climate zone
tropical_table <- wide_data %>%
  filter(climate_zone == "tropical") %>%
  select(-climate_zone)

# get directions of significant shifts
tropical_directions <- intersect(names(tropical_table), c("north", "south", "east", "west"))

# Output the tropical table
# tropical_table

table_3 <- tropical_table %>% 
  mutate(across(all_of(tropical_directions), str_to_sentence))

table_3_formatted <- knitr::kable(table_3)


# Filter for the subtropical climate zone
subtropical_table <- wide_data %>%
  filter(climate_zone == "subtropical") %>%
  select(-climate_zone)

# get directions of significant shifts
subtropical_directions <- intersect(names(subtropical_table), c("north", "south", "east", "west"))

table_4 <- subtropical_table %>% 
  mutate(across(all_of(subtropical_directions), str_to_sentence))

table_4_formatted <- knitr::kable(table_4)



# Filter for the temperate climate zone
temperate_table <- wide_data %>%
  filter(climate_zone == "temperate") %>%
  select(-climate_zone)

# get directions of significant shifts
temperate_directions <- intersect(names(temperate_table), c("north", "south", "east", "west"))

table_5 <- temperate_table %>% 
  mutate(across(all_of(temperate_directions), str_to_sentence))

table_5_formatted <- knitr::kable(table_5)


```

**Table XX.** The tropical fish species that have significantly shifted their geographical center of biomass (gCOB) over time.
`r table_3_formatted`

**Table XX.** The subtropical fish species that have significantly shifted their geographical center of biomass (gCOB) over time.
`r table_4_formatted`

**Table XX.** The temperate fish species that have significantly shifted their geographical center of biomass (gCOB) over time.
`r table_5_formatted`




```{r make figures of gCOB coefficient estimates}
#| echo: false

plot_lat_long_gCOB_data(lat_data = lat_models_df_yr2010_2022_m101112,
                        long_data = long_models_df_yr2010_2022_m101112,
                        title_prefix = "2010-2022 October, November, & December")


# should produce four figures
# Fig_1_yr2010_2022_m101112
# Fig_2_yr2010_2022_m101112
# Fig_3_yr2010_2022_m101112
# Fig_4_yr2010_2022_m101112


```


```{r Fig_2,  fig.width=8.5, fig.height=13}
#| echo: false

Fig_1_yr2010_2022_m101112

# fig.cap = "all species latitude gCOBs",

```

**Figure 2.** Latitudinal trends in gCOBs for fish species (deep-water, subtropical, temperate, tropical).




```{r Fig_3,  fig.width=8.5, fig.height=13}
#| echo: false

Fig_2_yr2010_2022_m101112

# fig.cap = "all species longitude gCOBs",

```

**Figure 3.**Longitudinal trends in gCOBs for fish species (deep-water, subtropical, temperate, tropical).



```{r Fig_4,  fig.width=8.5, fig.height=6.5}
#| echo: false

Fig_3_yr2010_2022_m101112

# fig.cap = "significant latitude gCOBs",

```

**Figure 4.**Latitudinal trends in gCOBs for fish species (deep-water, subtropical, temperate, tropical) with a significant shift over time.



```{r Fig_5,  fig.width=8.5, fig.height=5.5}
#| echo: false

Fig_4_yr2010_2022_m101112

# fig.cap = "significant longitude gCOBs",

```

**Figure 5.**Longitudinal trends in gCOBs for fish species (deep-water, subtropical, temperate, tropical) with a significant shift over time.




